---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# anvil

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
![R-CMD-check](https://github.com/r-xla/anvil/actions/workflows/R-CMD-check.yaml/badge.svg)
[![CRAN status](https://www.r-pkg.org/badges/version/anvil)](https://CRAN.R-project.org/package=anvil)
[![codecov](https://codecov.io/gh/r-xla/anvil/branch/main/graph/badge.svg)](https://codecov.io/gh/r-xla/anvil)
<!-- badges: end -->

Composable code transformation framework for R, allowing you to run numerical programs at the speed of light.
It currently implements JIT compilation for very fast execution and backward-mode automatic differentiation.
Programs can run on various hardware backends, including CPU and GPU.

## Installation

In order to install from source, you need a C++20 compiler, as well as `libprotobuf` and the `protobuf-compiler`.

``` r
pak::pak("r-xla/anvil")
```

You can also install from
[r-universe](https://r-xla.r-universe.dev/builds), by adding the code
below to your `.Rprofile`.

``` r
options(repos = c(
  rxla = "https://r-xla.r-universe.dev",
  CRAN = "https://cloud.r-project.org/"
))
```

## Quick Start

Below, we create a standard R function. We cannot directly call this function, but first need to wrap it in a `jit()` call.
If the resulting function is then called on `AnvilTensor`s -- the primary data type in {anvil} -- it will be JIT compiled and subsequently executed.

```{r}
library(anvil)
f <- function(a, b, x) {
  a * x + b
}
f_jit <- jit(f)

a <- nv_scalar(1.0)
b <- nv_scalar(-2.0)
x <- nv_scalar(3.0)

f_jit(a, b, x)
```

Through automatic differentiation, we can also obtain the gradient of the above function.

```{r}
g_jit <- jit(gradient(f, wrt = c("a", "b")))
g_jit(a, b, x)
```

## Main Features

- Automatic Differentiation:
  - Gradients for functions with scalar outputs are supported.
- Fast:
  - Code is JIT compiled into a single kernel.
  - Runs on different hardware backends, including CPU and GPU.
- Extendable:
  - It is possibel to add new primitives, transformations, and (with some effort) new backends.
  - The package is written almost entirely in R.

## When to use this package?

While {anvil} allows to run certain types of programs extremely fast, it only applies to a certain category of problems.
Specifically, it is suitable for numerical algorithms, such as optimizing bayesian models, training neural networks or more generally numerical optimization.
Another restriction is that {anvil} needs to re-compile the code for each new unique input shape.
This has the advantage, that the compiler can make memory optimizations, but the compilation overhead might be a problem for fast running programs.

## Acknowledgments

- This work is supported by [MaRDI](https://www.mardi4nfdi.de).
- The design of this package was inspired by and borrows from:
  - JAX, especially the [autodidax tutorial](https://docs.jax.dev/en/latest/autodidax.html).
  - The [microjax](https://github.com/joey00072/microjax) project.
- For JIT compilation, we leverage the [OpenXLA](https://openxla.org/) project.
