---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, you will learn everything you need to know to get started implementing numerical algorithms using {anvil}.
If you have experience with JAX in Python, you should feel right at home.

## The `AnvilArray`

We will start by introducing the main data structure, which is the `AnvilArray`.
It is essentially like an R array, with some differences:

1. It supports more data types, such as different precisions, as well as unsigned integers.
2. The array is managed by a specific backend (which we will ignore for now) and can live on different *device*s, such as CPU (aka host) or GPU.
3. 0-dimensional arrays can be used to represent scalars.

We can create an `AnvilArray` from R objects using `nv_array`.
Below, we create a 0-dimensional (i.e., a scalar) that holds a 16-bit integer on the CPU.

```{r}
library(anvil)
set.seed(42)
nv_array(1L, dtype = "i16", device = "cpu", shape = integer())
```

Note that for the creation of scalars, you can also use `nv_scalar` as a shorthand to skip specifying the shape and omit specifying the device, as CPU is the default.

```{r}
x <- nv_scalar(1L, dtype = "i16")
x
```

We can also create higher-dimensional arrays, for example a `2x3` array with single-precision floating-point numbers.
Without specifying the data type, it will default to `"f32"` for R doubles and `"i32"` for integers.

```{r}
y <- nv_array(1:6, shape = c(2, 3))
y
```

The `as_array()` function allows to convert `AnvilArray`s back to R objects.
Note that for 0-dimensional arrays, the result is an R vector of length 1, as R arrays cannot have 0 dimensions.

```{r}
as_array(y)
```

## Transforming Arrays

While the primary goal of anvil is to enable jit-compilation of R functions, it is also possible to directly transform `AnvilArray`s.
This is often referred to as *eager mode*, as opposed to *jit mode*.
There are two types of functions in {anvil}:

1. Anvil primitives, that follow the naming scheme `prim_<opname>`.
2. The main User API (`nv_<op>` functions).

The primitives define the fundamental operations that can be expressed in {anvil}.
But there are rather low-level and often lack some ergonomics such as type promotion (so we can add an `i32` to `f32`, see `vignette("type-promotion")` for details) or broadcacsting (so we can add a scalar to a `2x2` matrix):


```{r, error = TRUE}
prim_add(y, y)
prim_add(y, x)
nv_add(y, x)
```

More generally, R functions built out of `prim_*` operations (which includes the `nv_*` functions)  -- or the standard R operators that dispatch to them -- can be called directly on `AnvilArray`s.
For example, we can define a function that computes the output of a linear model $y = X \beta + \alpha$ and call it on anvil array.
We could have also used the overloaded `%*%` and `+` operators, but chose the underlying `nv_*` function for clarity.

```{r}
linear_model_r <- function(X, beta, alpha) {
  y0 <- nv_matmul(X, beta)
  nv_add(y0, alpha)
}

X <- nv_array(rnorm(6), dtype = "f32", shape = c(2, 3))
beta <- nv_array(rnorm(3), dtype = "f32", shape = c(3, 1))
alpha <- nv_scalar(rnorm(1), dtype = "f32")

linear_model_r(X, beta, alpha)
```

### Value Semantics

`AnvilArray`s are immutable and {anvil} follows **value semantics** -- exactly like R itself.
For example, assigning to a subset of an array inside a function does not affect the caller's array:

```{r}
modify <- function(x) {
  x[1, 1] <- 99
  x
}
y_new <- modify(y)
y       # unchanged
y_new   # modified copy
```

This would behave exactly the same way if `y` were an ordinary R array.
The performance implications, and how to opt into in-place reuse via the `donate` argument of `jit()`, are covered in the [Mutability and Donation](performance.html#mutability-and-donation) section of the performance vignette.

This eager execution is primarily intended for interactive exploration, prototyping and debugging.
The driving design goal for {anvil} is to enable `jit()` compiling complete functions.

## JIT Compilation

To get the most out of the XLA compiler, we can `jit()` a whole function so that it is compiled into an optimized executable.
For example, we can jit-compile the `linear_model_r` function we defined earlier:

```{r}
linear_model <- jit(linear_model_r)
linear_model(X, beta, alpha)
```

Note that `jit()` itself does not actually compile anything yet -- it just wraps `linear_model_r` in a closure.
The real compilation happens lazily on the first call, once the input shapes and data types are known; subsequent calls with the same input types reuse the cached executable.

The result is the same as calling `linear_model_r` directly.
The difference is that in eager mode, each operation is executed on its own, while `jit()` compiles the whole function into a single executable.
This removes per-call overhead and lets XLA optimize across the full computation, which typically makes `jit()`-compiled functions noticeably faster -- especially for larger computations or when the same function is called many times.

Under the hood, each `prim_*` function is itself just a `jit()`-compiled wrapper around a single primitive operation.

At the jit boundary, plain R values are transparently converted to `AnvilArray`s, so you don't need to wrap every input in `nv_array()` yourself.
This applies to both `nv_*` calls and your own `jit()`-compiled functions:

```{r}
nv_add(1, array(2:3))
```

For a more detailed discussion of when to prefer each mode, see `vignette("performance")`.

One restriction of {anvil} is that a function has to be re-compiled for every unique combination of input types, each consisting of a specific shape and data type.
To demonstrate this, we create a slightly modified version of our previous linear predictor function:

```{r}
linear_model2 <- jit(function(X, beta, alpha) {
  cat("compiling ...\n")
  X %*% beta + alpha
})
```

Next, we create a little helper function that creates example inputs with different numbers of observations:

```{r}
simul_data <- function(n, p) {
  list(
    X = nv_array(rnorm(n * p), dtype = "f32", shape = c(n, p)),
    beta = nv_array(rnorm(p), dtype = "f32", shape = c(p, 1)),
    alpha = nv_scalar(rnorm(1), dtype = "f32")
  )
}
```

Below, we call the function twice on data with the same shapes.

```{r}
do.call(linear_model2, simul_data(2, 3))
do.call(linear_model2, simul_data(2, 3))
```

We can notice that we only see the `"compiling ..."` message the first time, where the function is first compiled into an XLA executable, cached, and executed.
The second time, the executable is retrieved from the cache (because the inputs have the same shapes and data types) and executed without recompilation.
Because the executable contains only the operations applied to `AnvilArray`s, it does not contain the `cat()` call, so we don't see it the second time.

If we call the function on data with different shapes (or data types), the function is re-compiled and the message re-appears.

```{r}
y_hat <- do.call(linear_model2, simul_data(4, 3))
```

Because the compilation step itself can take some time, {anvil} therefore gives the best results when the same function is called many times with the same (or only a few different) input types, or the computation itself is sufficiently large to amortize the compilation overhead.

### Static Arguments

Besides `AnvilArray`s, jit-compiled functions can also take in regular R values as arguments.
For example, we might want a linear model with or without a bias term.
To do so, we add the `logical(1)` argument `with_bias` to our function.
We need to mark this argument as `static`, so {anvil} knows to treat it as a regular R value.

```{r}
linear_model3 <- jit(function(X, beta, alpha = NULL, with_bias) {
  if (with_bias) {
    cat("Compiling with bias ...\n")
    X %*% beta + alpha
  } else {
    cat("Compiling without bias ...\n")
    X %*% beta
  }
}, static = "with_bias")
```

We can now call this function with or without a bias term:

```{r}
linear_model3(X, beta, with_bias = FALSE)
linear_model3(X, beta, alpha, with_bias = TRUE)
```

Static arguments work differently than `AnvilArrays` as the function will be re-compiled for each new observed value of the static argument, not only each unique input type combination.

### Nested Inputs and Outputs

Inputs and outputs can also be nested data structures that contain `AnvilArray`s, although we currently only support (named) lists.

```{r}
linear_model4 <- jit(function(inputs) {
  list(y_hat = inputs[[1]] %*% inputs[[2]] + inputs[[3]])
})
linear_model4(list(X, beta, alpha))
```

### Composability

Jit-compiled functions can be called inside other `jit()` calls.
When this happens, the inner function is not compiled and executed separately -- instead, its operations are inlined into the outer computation, as if you had called the original R function directly.

To illustrate this, we jit-compile a mean squared error (MSE) loss function and check that it works standalone:

```{r}
mse <- jit(function(y_hat, y) {
  mean((y_hat - y)^2.0)
})

y <- nv_array(rnorm(2), dtype = "f32", shape = c(2, 1))
mse(linear_model(X, beta, alpha), y)
```

We now combine the jit-compiled `linear_model` and `mse` into a jitted `model_loss`:

```{r}
model_loss <- jit(function(X, beta, alpha, y) {
  y_hat <- linear_model(X, beta, alpha)
  mse(y_hat, y)
})

model_loss(X, beta, alpha, y)
```

Even though `linear_model` and `mse` are each jit-compiled, they are not run as separate executables inside `model_loss` -- their operations are folded into the outer graph so XLA can optimize the full computation as a whole.
Each piece can still be used standalone, where it is compiled and executed on its own.

So far, we have implemented the prediction step and the loss for the linear model.
One of the core applications of {anvil} is to implement learning algorithms, for which we often need gradients, as well as control flow.
We will start with gradients.

## Automatic Differentiation

In {anvil}, you can easily obtain the gradient function of a scalar-valued function via `gradient()`.
Currently, we don't support jacobians or hessians, but this will hopefilly be added in the future.

To fit the linear model, we first simulate some training data:

```{r}
beta <- rnorm(1)
X <- matrix(rnorm(100), ncol = 1)
alpha <- rnorm(1)
y <- X %*% beta + alpha + rnorm(100, sd = 0.5)
plot(X, y)
X <- nv_array(X)
y <- nv_array(y)
```

Next, we randomly initialize the model parameters:

```{r}
beta_hat <- nv_array(rnorm(1), shape = c(1, 1), dtype = "f32")
alpha_hat <- nv_scalar(rnorm(1), dtype = "f32")
```

Using the `gradient()` transformation, we can automatically obtain the gradient function of our `model_loss` from the composability section with respect to some of its arguments, which we specify.

```{r}
model_loss_grad <- gradient(
  model_loss,
  wrt = c("beta", "alpha")
)
```

Finally, we define the update step for the weights using gradient descent.

```{r}
update_weights <- jit(function(X, beta, alpha, y) {
  lr <- 0.1
  grads <- model_loss_grad(X, beta, alpha, y)
  beta_new <- beta - lr * grads$beta
  alpha_new <- alpha - lr * grads$alpha
  list(beta = beta_new, alpha = alpha_new)
})
```

This already allows us to fit the linear model

```{r}
weights <- list(beta = beta_hat, alpha = alpha_hat)
for (i in 1:100) {
  weights <- update_weights(X, weights$beta, weights$alpha, y)
}
```

```{r, echo = FALSE}
slope <- as_array(weights$beta)
intercept <- as_array(weights$alpha)
plot(as_array(X), as_array(y))
abline(as_array(weights$alpha), as_array(weights$beta), col = "red")
```

While this might seem like a reasonable solution, it continuously switches between the R interpreter and the XLA runtime. Moreover, we allocate new arrays in each iteration for the weights.
While the latter might not be a big problem for small models, it can cause significant overhead when working with bigger arrays.

Next, we will discuss control flow in {anvil} before we address immutability.

## Control Flow

In principle, there are three ways to implement control-flow in {anvil}:

1. Embed jit-compiled functions inside R control-flow constructs, which we have seen earlier.
2. Embed R control flow inside a jit-compiled function (we have also seen this earlier when our linear model allowed to optionally include a bias term).
3. Use special control-flow primitives provided by anvil, such as `nv_while()` and `nv_if()`.

Which solution is best depends on the specific scenario, so we will cover all three cases, at the risk of being a bit repetitive.
We will illustrate this with our linear model training example from earlier.
The first implementation is what we have already seen earlier: we jit-compile the update step and then repeatedly call it in an R loop:

```{r}
n_steps <- 100L
beta_hat <- nv_array(rnorm(1), shape = c(1, 1), dtype = "f32")
alpha_hat <- nv_scalar(rnorm(1), dtype = "f32")

weights <- list(beta = beta_hat, alpha = alpha_hat)
for (i in seq_len(n_steps)) {
  weights <- update_weights(X, weights$beta, weights$alpha, y)
}
weights
```

For simple update steps, this solution can be inefficient because every call into a jit-compiled function has some overhead.
How significant this overhead is depends on how expensive each call in the loop is -- for expensive functions the overhead becomes negligible.

The second approach is to use an R loop within the jit-compiled function.
There, the loop will be unrolled during the compilation step (for conditionals, only one branch is included in the executable as discussed earlier).
This will be rather slow in the example at hand, because we will also re-compute the gradient function in each iteration.
Moreover, the parameter `n_steps` is static, which means that for every unique value of `n_steps`, the function will be re-compiled into a different executable.

```{r}
train_unrolled <- jit(function(X, beta, alpha, y, n_steps) {
  lr <- nv_scalar(0.1)
  for (i in seq_len(n_steps)) {
    grads <- model_loss_grad(X, beta, alpha, y)
    beta <- beta - lr * grads$beta
    alpha <- alpha - lr * grads$alpha
  }
  list(beta = beta, alpha = alpha)
}, static = "n_steps")

beta_hat <- nv_array(rnorm(1), shape = c(1, 1), dtype = "f32")
alpha_hat <- nv_scalar(rnorm(1), dtype = "f32")
train_unrolled(X, beta_hat, alpha_hat, y, n_steps = 10L)
```

Finally, the third approach is to use the `nv_while` function.
It is not like a standard while loop, because `anvil` is purely functional.

The function takes in:

1. An initial state, which is a (nested) list of `AnvilArray`s.
1. A `cond` function, which takes as input the current state and returns a logical flag indicating whether to continue the loop.
1. A `body` function, which takes as input the current state and returns a new state.

```{r}
train_while <- jit(function(X, beta, alpha, y, n_steps) {
  lr <- 0.1
  nv_while(
    list(beta = beta, alpha = alpha, i = nv_scalar(0L)),
    \(beta, alpha, i) i < n_steps,
    \(beta, alpha, i) {
      grads <- model_loss_grad(X, beta, alpha, y)
      list(
        beta = beta - lr * grads$beta,
        alpha = alpha - lr * grads$alpha,
        i = i + 1L
      )
    }
  )
})

beta_hat <- nv_array(rnorm(1), shape = c(1, 1), dtype = "f32")
alpha_hat <- nv_scalar(rnorm(1), dtype = "f32")
train_while(X, beta_hat, alpha_hat, y, nv_scalar(100L))
```

The same approach works analogously for `if`-statements, where the {anvil} primitive `nv_if` is available.

## Donating Arguments

Because `AnvilArray`s are immutable, every call to `update_weights` above allocates a fresh buffer for the updated weights, and the previous `beta`/`alpha` arrays become garbage.
When you know the inputs are no longer needed after a call -- as is the case for weight updates -- you can mark them as `donate` so their memory can be reused instead:

```r
update_weights_donatable <- jit(update_weights_fn, donate = c("beta", "alpha"))
```

After such a call, the original arrays are unusable (reading from them errors), but XLA is free to perform the update in place.
See the [Mutability and Donation](performance.html#mutability-and-donation) section of the performance vignette for a detailed discussion and a worked example.
