---
title: "Correct Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correct Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to write functions that produce the same, correct results both when called eagerly and when wrapped in `jit()`.
Eager and traced execution have different semantics -- under `jit()`, arrayish inputs become `GraphBox` tracers that do not expose their concrete values -- so some care is required to keep a function correct in both modes.
For general background on how `jit()` traces your code and the pitfalls that come with tracing-based compilation, see the [JIT Deep Dive](jit.html) vignette.
The `nv_*` functions shipped with {anvl} (e.g. `nv_add()`, `nv_matmul()`) follow the patterns below, and you should too.

A correct function:

1. Accepts the standard range of arrayish inputs: length-1 R vectors, R arrays, and `AnvlArray`s.
2. Always returns an `AnvlArray`.
3. Works correctly both eagerly and under `jit()`.

```{r setup}
library(anvl)
```

## Input conversion

anvl functions typically accept R literals (length-1 vectors, R arrays), `AnvlArray`s, and (during tracing) graph boxes.
Primitives handle all of these, but having the same arrayish value arrive as three different R types makes the function body awkward: every branch would have to re-check which case it is in.
To keep the body simple, we standardize at the top -- after that, we can treat every argument as an `AnvlArray` (or a tracing box) without caring about the original R type.
Anvl provides two helpers for this:

* `as_anvl_array(x, device = NULL)` -- use for functions that take a **single** array input.
  It converts an R value to an `AnvlArray` on the given device (or the default device if `NULL`), and passes `AnvlArray`s through.
* `as_anvl_arrays(...)` -- use for functions with **multiple** array inputs.
  It infers a common device from any concrete `AnvlArray` in `...`, places R literals on that device, and errors if the concrete inputs live on different devices (or backends).

Both helpers short-circuit during tracing: inside `jit()`, device placement is already handled at the input/output boundary and when boxing constants.

### Single input

```{r}
nv_expand <- function(x, shape) {
  x <- as_anvl_array(x)
  if (identical(shape(x), shape)) {
    return(x)
  }
  nv_broadcast_to(x, shape)
}
```

### Multiple inputs

Use `as_anvl_arrays()` to normalize all arrayish inputs in one call:

```{r}
nv_add_scaled <- function(x, y, alpha) {
  args <- as_anvl_arrays(x, y, alpha)
  args[[1L]] + args[[2L]] * args[[3L]]
}
```

## Always output anvl arrays

As long as you normalize the inputs with `as_anvl_array()` / `as_anvl_arrays()` and then build the result from primitives (or from values that already went through that conversion), the output is automatically an `AnvlArray`.
The one thing to watch for is short-circuit returns: if a code path returns one of the inputs unchanged, make sure the conversion step happened first so you are returning an `AnvlArray` and not the original R value.

## Creating constants on the right device

If your function creates a constant inside its body (via `nv_fill()`, `nv_iota()`, `nv_scalar()`, `nv_eye()`, ...), the constant must live on the same backend/device as the input.
The idiomatic way to do this is with the `nv_<op>_like()` variants, which default `dtype`, `shape`, `backend`, `device`, and `ambiguous` from an example input:

```{r}
nv_tril <- function(x) {
  nv_ifelse(x > 0, x, nv_fill_like(x, 0))
}
jit(nv_tril)(nv_array(matrix(c(-1, 2, -3, 4), 2, 2)))
```

If you write your own array-creator function, consider providing a matching `_like()` variant so callers can follow the same pattern.

## Inputs that require checks must be static

Any argument whose value the function inspects (e.g. for validation, to pick a code path, or to compute a shape or `dim` index) must be a **static** argument under `jit()`.
Under tracing, arrayish inputs are replaced by `GraphBox` tracers and their concrete R value is not available -- assertions on the value would either fail or see the tracer instead of the user's data.
Static arguments, in contrast, are embedded as constants into the compiled program, so the function body sees their real R value during tracing.

As a rule of thumb, if your function body contains `if`, `switch`, `stopifnot`, `checkmate::assert_*`, or any other code that branches on or validates a value, that argument should be static.
This typically applies to arguments like `dims`, `shape`, `dim`, flags, mode strings, or dtype specifiers -- anything that controls the *structure* of the computation rather than participating in it as data.

```r
nv_reduce_sum <- function(x, dims = NULL) {
  if (!is.null(dims)) {
    assert_integerish(dims, lower = 1L, upper = ndims(x))
  }
  # ...
}

# `dims` is inspected, so it must be static when jitted:
jit(nv_reduce_sum, static = "dims")(x, dims = 1L)
```

Arrayish inputs (the actual data being operated on) should *not* be static -- they are meant to be traced.

## Binary element-wise operations

Element-wise binary ops (like `nv_add()`, `nv_mul()`, comparisons) share a fixed preparation pipeline: promote both operands to a common dtype via `nv_promote_to_common()`, then broadcast a scalar operand up to the non-scalar shape via `nv_broadcast_scalars()`, and finally dispatch to the primitive:

```r
nv_add <- function(lhs, rhs) {
  args <- nv_promote_to_common(lhs, rhs)
  args <- nv_broadcast_scalars(args[[1L]], args[[2L]])
  prim_add(args[[1L]], args[[2L]])
}
```

If instead you need full NumPy-style broadcasting (not just scalar-against-tensor -- e.g. for an outer product), use `nv_broadcast_arrays()` after promotion.
