---
title: "Differences from base R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differences from base R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette lists major behavioral differences between {anvl} and base R that R users should be aware of when working with `AnvlArray`s.

```{r}
library(anvl)
```

## Row-major vs column-major ordering

R stores matrices and arrays in *column-major* order, while {anvl} (following XLA) uses *row-major* order.
This makes no difference when you only use shape-aware operations (subsetting, matrix multiplication, etc.) -- the indices are the same in both.
The difference shows up when you flatten an array, because the underlying data is then traversed in a different order.

Consider the 2x2 matrix below:

```{r}
m <- matrix(1:4, nrow = 2)
m
```

In base R, `as.vector()` flattens it column-by-column, so we get `1, 2, 3, 4`:

```{r}
as.vector(m)
```

In {anvl}, reshaping to a length-4 vector traverses the data row-by-row, so we get `1, 3, 2, 4`:

```{r}
nv_reshape(m, shape = 4)
```

If you need column-major flattening in {anvl}, transpose first:

```{r}
nv_reshape(nv_transpose(m), shape = 4)
```

## No recycling

Base R *recycles* the shorter operand when two vectors of different lengths are combined elementwise:

```{r}
c(1, 2, 3, 4) + c(1, 2)
```

{anvl} only auto-broadcasts *scalars* (operands with shape `integer()`).
Adding a scalar to an array works as you would expect:

```{r}
nv_array(1:4) + 10L
```

But combining two non-scalar arrays of different shapes errors, even when one shape is a "tile" of the other:

```{r, error = TRUE}
nv_array(1:4) + nv_array(1:2)
```

When two non-scalar arrays differ only by size-1 dimensions (numpy-style broadcasting, e.g. shape `(2, 3)` and `(1, 3)`), use `nv_broadcast_arrays()` to align them explicitly first:

```{r}
a <- nv_array(matrix(1:6, nrow = 2))
b <- nv_array(matrix(c(10, 20, 30), nrow = 1))
xs <- nv_broadcast_arrays(a, b)
xs[[1]] + xs[[2]]
```

Note that even `nv_broadcast_arrays()` cannot replicate R's recycling for shapes like `(4)` and `(2)` -- the shapes must be broadcast-compatible in the numpy sense.

## No `NA`s

R has a dedicated missing-value marker (`NA`) for every atomic type.
{anvl} arrays do not -- there is no representation of "missing" at the XLA level.
When you convert R values containing `NA` into an `AnvlArray`, the `NA`s are silently turned into the closest available value of the target dtype.
For floating-point dtypes, that value is `NaN`:

```{r}
nv_array(NA_real_)
```

```{r}
nv_array(c(1, NA, 3))
```

Round-tripping back to R produces `NaN`, not `NA`:

```{r}
as_array(nv_array(c(1, NA, 3)))
```

Integer dtypes have no `NaN`, but `NA_integer_` does *appear* to round-trip:

```{r}
nv_scalar(NA_integer_) |> as.integer()
```

This is misleading.
R represents `NA_integer_` by reserving one specific 32-bit integer value (`-2^31 = -2147483648`) as a sentinel for missingness, leaving only `2^32 - 1` valid integers.
{anvl} has no notion of missing values and just stores that bit pattern as a regular `i32`.
The round-trip "works" only because R interprets the same bit pattern back as `NA` -- but inside {anvl} the value behaves like the integer `-2147483648`, and any computation on it (e.g. addition, comparison) will treat it as such rather than propagating missingness.
The same caveat applies in reverse: if a genuine {anvl} computation produces `-2147483648`, converting back to R will silently turn it into `NA`.

For logical (`bool`) dtype the situation is worst: there is no spare bit pattern at all, so a bare `NA` (which is logical) is silently turned into `TRUE`:

```{r}
nv_scalar(NA)
as.logical(nv_scalar(NA))
```

If your data contains missing values, decide how to handle them *before* converting to an `AnvlArray`.
