---
title: "Gaussian Process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian Process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we implement a Gaussian Process (GP) regression model from scratch.

## Background

A Gaussian Process is a distribution over functions, fully specified by a mean function $m(x)$ and
a covariance (kernel) function $k(x, x')$.
Given $n$ training observations $\mathbf{y}$ at inputs $\mathbf{X}$, the GP posterior predictive
distribution at new inputs $\mathbf{X}_*$ is:

$$\mathbf{f}_* | \mathbf{X}, \mathbf{y}, \mathbf{X}_* \sim \mathcal{N}(\boldsymbol{\mu}_*, \boldsymbol{\Sigma}_*)$$

where

$$\boldsymbol{\mu}_* = K(\mathbf{X}_*, \mathbf{X}) \, K(\mathbf{X}, \mathbf{X})^{-1} \, \mathbf{y}$$
$$\boldsymbol{\Sigma}_* = K(\mathbf{X}_*, \mathbf{X}_*) - K(\mathbf{X}_*, \mathbf{X}) \, K(\mathbf{X}, \mathbf{X})^{-1} \, K(\mathbf{X}, \mathbf{X}_*)$$

We will use the squared exponential (RBF) kernel:

$$k(x, x') = \sigma_f^2 \exp\!\left(-\frac{(x - x')^2}{2 \ell^2}\right)$$

with lengthscale $\ell$ and signal variance $\sigma_f^2$.

## Data

We generate synthetic data from a noisy sinusoidal function.

```{r}
library(anvil)
set.seed(42)

n_train <- 20
noise_sd <- 0.2

X_train <- sort(runif(n_train, -5, 5))
y_train <- sin(X_train) + rnorm(n_train, sd = noise_sd)

n_test <- 100
X_test <- seq(-6, 6, length.out = n_test)

plot(X_train, y_train, pch = 19, xlab = "x", ylab = "y", main = "Training Data")
curve(sin, from = -6, to = 6, add = TRUE, col = "blue", lwd = 2)
legend("topright", legend = c("Observations", "True function"),
  pch = c(19, NA), lty = c(NA, 1), col = c("black", "blue"), lwd = c(NA, 2))
```

## Kernel

The squared exponential kernel computes pairwise distances between input points and applies the RBF
formula.
We parameterize it with log-transformed hyperparameters for unconstrained optimization.

```{r}
rbf_kernel <- function(X1, X2, log_lengthscale, log_signal_var) {
  lengthscale <- exp(log_lengthscale)
  signal_var <- exp(log_signal_var)

  # Compute squared distances: ||x - x'||^2 = ||x||^2 - 2 x^T x' + ||x'||^2
  # X1 is (n, d), X2 is (m, d)
  sq1 <- nv_reduce_sum(X1 * X1, dims = 2L, drop = FALSE) # (n, 1)
  sq2 <- nv_reduce_sum(X2 * X2, dims = 2L, drop = FALSE) # (m, 1)
  cross <- X1 %*% t(X2) # (n, m)
  # Broadcast: (n,1) and (1,m) to (n,m)
  sq_dist <- nv_broadcast_to(sq1, shape(cross)) -
    2 * cross +
    nv_broadcast_to(t(sq2), shape(cross))

  signal_var * exp(-sq_dist / (2 * lengthscale^2))
}
```

## Marginal Likelihood

The log marginal likelihood of the GP is:

$$\log p(\mathbf{y} | \mathbf{X}, \boldsymbol{\theta}) = -\frac{1}{2} \mathbf{y}^\top K_y^{-1} \mathbf{y} - \frac{1}{2} \log |K_y| - \frac{n}{2} \log 2\pi$$

where $K_y = K(\mathbf{X}, \mathbf{X}) + \sigma_n^2 I$ includes the noise variance.

Using the Cholesky decomposition $L = \text{chol}(K_y)$, we can compute this efficiently:

- $K_y^{-1} \mathbf{y} = L^{-\top} L^{-1} \mathbf{y}$ (two triangular solves via `nv_solve`)
- $\log |K_y| = 2 \sum_i \log L_{ii}$

```{r}
neg_log_marginal_likelihood <- function(X, y, log_lengthscale, log_signal_var, log_noise_var) {
  n <- shape_abstract(y)[1L]
  noise_var <- exp(log_noise_var)

  K <- rbf_kernel(X, X, log_lengthscale, log_signal_var)
  # Add noise variance to diagonal: K_y = K + sigma_n^2 * I
  noise_diag <- noise_var * nv_tensor(diag(n), dtype = dtype_abstract(K))
  K_y <- K + noise_diag

  L <- nvl_cholesky(K_y, lower = TRUE)

  # alpha = K_y^{-1} y = solve(K_y, y)
  alpha <- nv_solve(K_y, y)

  # Data fit term: 0.5 * y^T @ alpha
  data_fit <- 0.5 * nv_reduce_sum(y * alpha, dims = c(1L, 2L))

  # Log determinant: sum(log(diag(L)))
  # Extract diagonal using a mask
  diag_mask <- nv_tensor(diag(n), dtype = dtype_abstract(L))
  diag_L <- nv_reduce_sum(L * diag_mask, dims = 2L)
  log_det <- nv_reduce_sum(log(diag_L), dims = 1L)

  # Constant term
  const <- 0.5 * n * log(2 * pi)

  data_fit + log_det + const
}
```

## Hyperparameter Optimization

We optimize the kernel hyperparameters by minimizing the negative log marginal likelihood using
gradient descent.

```{r}
X_t <- nv_tensor(matrix(X_train, ncol = 1), dtype = "f32")
y_t <- nv_tensor(matrix(y_train, ncol = 1), dtype = "f32")

params <- c("log_lengthscale", "log_signal_var", "log_noise_var")
nll_grad <- gradient(neg_log_marginal_likelihood, wrt = params)

# Training loop
train_gp <- jit(function(X, y, log_ls, log_sv, log_nv, n_steps, lr) {
  nv_while(
    list(log_ls = log_ls, log_sv = log_sv, log_nv = log_nv, step = nv_scalar(0L)),
    \(log_ls, log_sv, log_nv, step) step < n_steps,
    \(log_ls, log_sv, log_nv, step) {
      grads <- nll_grad(X, y, log_ls, log_sv, log_nv)
      list(
        log_ls = log_ls - lr * grads$log_lengthscale,
        log_sv = log_sv - lr * grads$log_signal_var,
        log_nv = log_nv - lr * grads$log_noise_var,
        step = step + 1L
      )
    }
  )
})
```

```{r}
result <- train_gp(
  X_t, y_t,
  log_ls = nv_scalar(0, dtype = "f32"),
  log_sv = nv_scalar(0, dtype = "f32"),
  log_nv = nv_scalar(-1, dtype = "f32"),
  n_steps = nv_scalar(500L),
  lr = nv_scalar(0.01)
)

learned_ls <- as_array(result$log_ls)
learned_sv <- as_array(result$log_sv)
learned_nv <- as_array(result$log_nv)

cat(sprintf(
  "Learned hyperparameters:\n  lengthscale = %.3f\n  signal_var  = %.3f\n  noise_var   = %.3f\n",
  exp(learned_ls), exp(learned_sv), exp(learned_nv)
))
```

## Prediction

With the optimized hyperparameters, we compute the posterior predictive distribution.

```{r}
predict_gp <- jit(function(X_train, y_train, X_test, log_ls, log_sv, log_nv) {
  noise_var <- exp(log_nv)
  n <- shape_abstract(X_train)[1L]

  # Training kernel + noise
  K <- rbf_kernel(X_train, X_train, log_ls, log_sv)
  K_y <- K + noise_var * nv_tensor(diag(n), dtype = dtype_abstract(K))

  # Cross-covariance and test kernel
  K_s <- rbf_kernel(X_test, X_train, log_ls, log_sv)
  K_ss <- rbf_kernel(X_test, X_test, log_ls, log_sv)

  # Predictive mean: K_s @ K_y^{-1} @ y
  alpha <- nv_solve(K_y, y_train)
  mu <- K_s %*% alpha

  # Predictive covariance: K_ss - K_s @ K_y^{-1} @ K_s^T
  v <- nv_solve(K_y, t(K_s))
  cov <- K_ss - K_s %*% v

  list(mu = mu, cov = cov)
})

X_test_t <- nv_tensor(matrix(X_test, ncol = 1), dtype = "f32")

pred <- predict_gp(
  X_t, y_t, X_test_t,
  result$log_ls, result$log_sv, result$log_nv
)

mu <- as.vector(as_array(pred$mu))
var <- diag(as_array(pred$cov))
sd <- sqrt(pmax(var, 0))
```

## Results

```{r, fig.width=7, fig.height=5}
plot(X_test, mu, type = "l", lwd = 2, col = "red",
  ylim = range(c(mu - 2 * sd, mu + 2 * sd, y_train)),
  xlab = "x", ylab = "y", main = "Gaussian Process Regression")
polygon(
  c(X_test, rev(X_test)),
  c(mu + 2 * sd, rev(mu - 2 * sd)),
  col = rgb(1, 0, 0, 0.15), border = NA
)
lines(X_test, mu, lwd = 2, col = "red")
points(X_train, y_train, pch = 19)
curve(sin, from = -6, to = 6, add = TRUE, col = "blue", lwd = 2, lty = 2)
legend("topright",
  legend = c("GP mean", "95% interval", "Observations", "True function"),
  lty = c(1, NA, NA, 2), lwd = c(2, NA, NA, 2), pch = c(NA, 15, 19, NA),
  col = c("red", rgb(1, 0, 0, 0.3), "black", "blue"))
```

The GP successfully recovers the underlying sinusoidal function with well-calibrated uncertainty
bands that widen in regions far from training data.
