---
title: "Hamiltonian Monte Carlo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hamiltonian Monte Carlo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

In this vignette, we implement Hamiltonian Monte Carlo (HMC) [@neal2011mcmc] using {anvil}.
HMC augments a random-walk Metropolis-Hastings sampler with gradient information to propose large, well-informed moves across the target distribution, achieving much lower autocorrelation than plain MH.
Because HMC needs gradients of the log-density, it is a natural showcase for {anvil}'s three core code transformations working together: `jit()`, `gradient()`, and `nv_while()`.

We target the same "banana" distribution as the [Metropolis-Hastings vignette](metropolis-hastings.html) so the two samplers can be compared directly.

We write the sampler in a *batched* form from the start: every array carries a leading chain axis of size $K$, and the $K$ chains evolve independently.
A single chain is simply the $K = 1$ special case, while running many chains in parallel requires no additional code --- we just pass a $(K, 2)$ starting tensor with $K > 1$.
This mirrors how JAX / NumPyro use `vmap(grad(U))` internally; {anvil} does not yet expose `vmap`, but for HMC writing the kernel batched by hand is straightforward and pays off once we want to saturate a GPU.

For the Stan runtime comparison later in this vignette we want both samplers to run under matched conditions, so we pin every parallelism knob we can reach from R to a single thread *before* loading either framework.
For XLA this means `PJRT_NPROC=1` (which caps the PjRt CPU client's intra-op thread pool, [read on startup from the environment](https://github.com/openxla/xla/blob/main/xla/pjrt/utils.cc) and immutable afterwards) plus the debug flag `--xla_cpu_multi_thread_eigen=false` (which disables Eigen's within-op parallelism in generated code).
`OMP_NUM_THREADS=1` is already R's default for Rscript, so OpenMP-backed numerical libraries are single-threaded regardless.
For Stan we set `STAN_NUM_THREADS=1`; this only matters if the model uses `reduce_sum(...)` (ours does not), but we set it for symmetry.

```{r}
Sys.setenv(
  PJRT_NPROC = "1",
  XLA_FLAGS = "--xla_cpu_multi_thread_eigen=false",
  STAN_NUM_THREADS = "1"
)
library(anvil)
b_param <- 0.01
b_t <- nv_scalar(b_param, dtype = "f64")
```

## Potential Energy

HMC draws samples from $p(\theta)$ by simulating the dynamics of a particle moving in a "potential energy" landscape $U(\theta) = -\log p(\theta)$.
Unlike MH, which only queries the density at single points, HMC follows gradient flows of $U$ to move coherently across the distribution.

For the banana, dropping additive constants,

$$U(\theta_1, \theta_2) = \frac{\theta_1^2}{200} + \frac{(\theta_2 - b\theta_1^2 + 100b)^2}{2}.$$

We implement the potential as a function that takes a $(K, 2)$ parameter tensor --- one row per chain --- and returns a $K$-vector of per-chain energies.
For $K = 1$ this is just a length-1 vector.

```{r}
U <- function(theta, b) {
  theta1 <- theta[, 1]
  theta2 <- theta[, 2]
  theta1^2 / 200 + (theta2 - b * theta1^2 + 100 * b)^2 / 2
}
```

`gradient()` requires a scalar output, so to obtain per-chain gradients we take the gradient of $\sum_k U(\theta_k)$.
Because chain $k$'s energy only depends on $\theta_k$, the cross-chain entries of the Jacobian vanish and row $k$ of the returned gradient is exactly $\nabla U(\theta_k)$.
The $K$ chains' gradients are therefore computed in a single call, which is the same reduction trick `jax.vmap` applies internally to `grad(U)`.

```{r}
U_total <- function(theta, b) sum(U(theta, b))
grad_U <- gradient(U_total, wrt = "theta")
```

## Leapfrog Integrator

HMC introduces an auxiliary momentum variable $p$ with kinetic energy $K(p) = \tfrac{1}{2}\|p\|^2$ and simulates Hamiltonian dynamics on $H(\theta, p) = U(\theta) + K(p)$.
In continuous time these dynamics exactly preserve $H$; in discrete time we need an integrator that approximately preserves $H$ and is *reversible* and *volume-preserving*, so that a Metropolis accept/reject step still yields a valid MCMC sampler.

The standard choice is the **leapfrog integrator** (a.k.a. velocity Verlet), which alternates half-steps on the momentum and full steps on the position:

$$
\begin{aligned}
p &\gets p - \tfrac{\varepsilon}{2}\,\nabla U(\theta) &&\text{(half kick)} \\
\theta &\gets \theta + \varepsilon\,p &&\text{(drift)} \\
p &\gets p - \tfrac{\varepsilon}{2}\,\nabla U(\theta) &&\text{(half kick)}.
\end{aligned}
$$

We repeat this elementary step $L$ times.
A careful implementation evaluates the gradient exactly once per step by carrying $\nabla U(\theta)$ through the loop state as a cached value: the trailing half-kick of step $n$ uses the same gradient as the leading half-kick of step $n+1$, so reusing it avoids a redundant gradient evaluation.
This is the structure used by JAX-based samplers such as NumPyro, and maps cleanly onto `nv_while`.
Note that the return value of a function produced by `gradient()` is a list keyed by `wrt` names, so we access the gradient as `grad_U(theta, b)$theta`.

Every operation inside the loop is element-wise on the leading chain axis, so the same code handles any $K$.

```{r}
leapfrog <- function(theta, p, b, eps, L) {
  grad <- grad_U(theta, b)$theta
  out <- nv_while(
    list(theta = theta, p = p, grad = grad, i = nv_scalar(0L)),
    \(theta, p, grad, i) i < L,
    \(theta, p, grad, i) {
      p <- p - 0.5 * eps * grad
      theta <- theta + eps * p
      grad <- grad_U(theta, b)$theta
      p <- p - 0.5 * eps * grad
      list(theta = theta, p = p, grad = grad, i = i + 1L)
    }
  )
  list(theta = out$theta, p = out$p)
}
```

Each leapfrog trajectory of $L$ steps costs $L + 1$ gradient evaluations (shared across all chains): one before the loop, and one per iteration.

## The HMC Step

One HMC transition consists of:

1. Sampling a fresh momentum $p \sim \mathcal{N}(0, I)$ independently per chain.
2. Simulating an $L$-step leapfrog trajectory from $(\theta, p)$ to $(\theta', p')$ for every chain.
3. Accepting each chain's proposal $\theta'_k$ with its own probability $\min(1, \exp(H(\theta_k, p_k) - H(\theta'_k, p'_k)))$.

The accept/reject step corrects for the discretisation error introduced by leapfrog; in the limit $\varepsilon \to 0$ the Hamiltonian is preserved exactly and acceptance goes to 1.
As in the MH vignette, the accept/reject decision uses `nv_ifelse()` (not an R `if`), and the RNG state is threaded through explicitly.

We sum the squared momentum only along the parameter dimension (`dims = 2L`) so the kinetic energy is a $K$-vector, and we make the accept/reject decision per chain.
`nv_ifelse()` requires its predicate to have the same shape as the selected arrays, so we broadcast the length-$K$ accept vector up to $(K, 2)$:

```{r}
hmc_step <- function(theta, rng_state, b, eps, L) {
  rng_out <- nv_rnorm(shape(theta), rng_state, dtype = "f64")
  rng_state <- rng_out[[1L]]
  p <- rng_out[[2L]]

  H_current <- U(theta, b) + nv_reduce_sum(p^2, dims = 2L) / 2
  lf <- leapfrog(theta, p, b, eps, L)
  H_proposed <- U(lf$theta, b) + nv_reduce_sum(lf$p^2, dims = 2L) / 2
  log_accept <- H_current - H_proposed

  n_chains <- shape(theta)[1L]
  rng_out <- nv_runif(n_chains, rng_state, dtype = "f64")
  rng_state <- rng_out[[1L]]
  u <- rng_out[[2L]]

  accept <- log(u) < log_accept
  accept_2d <- nv_broadcast_to(nv_reshape(accept, c(n_chains, 1L)), shape(theta))
  new_theta <- nv_ifelse(accept_2d, lf$theta, theta)
  list(theta = new_theta, rng_state = rng_state)
}
```

We JIT-compile a single HMC step.
In contrast to MH, we do not thin the chain: a well-tuned HMC proposal moves far enough that successive samples are already approximately independent.

```{r}
hmc_sample <- jit(hmc_step)
```

## Running a Single Chain

Step size $\varepsilon$ and trajectory length $L$ are the two hyperparameters of HMC.
$\varepsilon$ must be small enough for leapfrog to be stable on the tightest direction of the target; $L \cdot \varepsilon$ should be roughly the length scale of the broadest direction.
For the banana, $\theta_1$ has standard deviation $10$ and $\theta_2$ has standard deviation close to $1.7$.
A trajectory that is too long will "U-turn" and land close to the starting point, producing anti-correlated samples; a trajectory that is too short behaves like plain random-walk MH.
A reasonable target is roughly a quarter-period of oscillation in the widest direction, i.e. $L \cdot \varepsilon \approx \tfrac{\pi}{2}\,\sigma_{\max} \approx 16$.
We pick $\varepsilon = 0.2$ (safely stable in the tight $\theta_2$ direction) and $L = 80$.

We start with a single chain --- the $K = 1$ special case of the batched kernel --- represented as a $(1, 2)$ tensor.

```{r}
n_samples <- 5000L
n_warmup <- 500L
eps <- nv_scalar(0.2, dtype = "f64")
L <- nv_scalar(80L)

theta <- nv_array(matrix(0, nrow = 1L, ncol = 2L), dtype = "f64")
rng_state <- nv_rng_state(42L)

for (i in seq_len(n_warmup)) {
  res <- hmc_sample(theta, rng_state, b_t, eps, L)
  theta <- res$theta
  rng_state <- res$rng_state
}

thetas <- vector("list", n_samples)
for (i in seq_len(n_samples)) {
  res <- hmc_sample(theta, rng_state, b_t, eps, L)
  theta <- res$theta
  rng_state <- res$rng_state
  thetas[[i]] <- theta
}

samples <- do.call(rbind, lapply(thetas, function(x) as.vector(as_array(x))))
colnames(samples) <- c("theta1", "theta2")
```

## Inspecting the Results

The autocorrelation function shows that HMC samples are much less correlated than plain MH samples: we did not need to thin at all.

```{r, echo = FALSE, fig.width = 7, fig.height = 4}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
acf(samples[, 1], main = expression(theta[1]), lag.max = 40)
acf(samples[, 2], main = expression(theta[2]), lag.max = 40)
```

Comparison against the analytical moments of the banana distribution:

```{r, echo = FALSE}
sample_mean <- colMeans(samples)
sample_sd <- apply(samples, 2, sd)

true_mean <- c(0, 0)
true_sd <- c(10, sqrt(2e4 * b_param^2 + 1))

comparison <- data.frame(
  Parameter = c("theta1", "theta2"),
  HMC_Mean = sample_mean,
  True_Mean = true_mean,
  HMC_SD = sample_sd,
  True_SD = true_sd,
  row.names = NULL
)
comparison
```

And the samples overlaid on the true density contours:

```{r, echo = FALSE, fig.width = 7, fig.height = 5}
log_p <- function(x1, x2) {
  -x1^2 / 200 - (x2 - b_param * x1^2 + 100 * b_param)^2 / 2
}
x1_grid <- seq(-30, 30, length.out = 200)
x2_grid <- seq(-10, 15, length.out = 200)
z <- outer(x1_grid, x2_grid, log_p)

par(mar = c(4, 4, 2, 1))
contour(
  x1_grid, x2_grid, exp(z), nlevels = 15,
  xlab = expression(theta[1]), ylab = expression(theta[2]),
  main = "HMC Samples vs True Density", col = "grey60"
)
points(
  samples[, 1], samples[, 2],
  pch = 16, cex = 0.3, col = adjustcolor("steelblue", 0.4)
)
```

## Moving the Sampling Loop into the Graph

The version of the sampler above calls a JIT-compiled function `hmc_sample()` once per iteration from an R `for` loop.
Each call pays a fixed cost for preparing the call, dispatching it into the XLA executable, and copying the result back out.
When the target distribution is large (e.g. a deep neural network's log-posterior) this overhead is negligible, but for a two-dimensional target it *dominates* the per-iteration cost of the actual leapfrog computation.

We can eliminate the per-iteration overhead entirely by lifting the outer sampling loop into the graph as well --- so that a single JIT-compiled function produces *all* samples in one executable call.
The main ingredient is an `nv_while` whose state carries, in addition to $\theta$ and the RNG, a pre-allocated output buffer into which successive samples are written.
{anvil}'s subset-assignment syntax (`samples[i, , ] <- value`) works inside traced functions, returning a new tensor with the slice updated, so this looks essentially like writing to a pre-allocated R array.

The buffer has shape $(n, K, 2)$ --- one slice per sample, per chain.
Because the shape must be known at compile time, we pass the number of samples `n` as a static argument; the chain count $K$ is read from the starting tensor's shape.

```{r}
hmc_sample_many <- jit(function(theta, rng_state, b, eps, L, n) {
  samples0 <- nv_fill(0, shape = c(n, shape(theta)[1L], 2L), dtype = "f64")
  out <- nv_while(
    list(theta = theta, rng_state = rng_state, samples = samples0, i = nv_scalar(0L)),
    \(theta, rng_state, samples, i) i < n,
    \(theta, rng_state, samples, i) {
      step <- hmc_step(theta, rng_state, b, eps, L)
      samples[i + 1L, , ] <- step$theta
      list(theta = step$theta, rng_state = step$rng_state, samples = samples, i = i + 1L)
    }
  )
  list(samples = out$samples, theta = out$theta, rng_state = out$rng_state)
}, static = "n")
```

The first call triggers JIT compilation, so we warm the cache once before timing the real run.

```{r}
theta <- nv_array(matrix(0, nrow = 1L, ncol = 2L), dtype = "f64")
rng_state <- nv_rng_state(42L)

warm <- hmc_sample_many(theta, rng_state, b_t, eps, L, n = n_warmup)
theta <- warm$theta
rng_state <- warm$rng_state
# Second call with the real n compiles a fresh executable (n is static)
discard <- hmc_sample_many(theta, rng_state, b_t, eps, L, n = n_samples)
theta <- discard$theta
rng_state <- discard$rng_state

t_anvil <- system.time({
  out <- hmc_sample_many(theta, rng_state, b_t, eps, L, n = n_samples)
  as_array(out$samples)  # force XLA to finish computing before stopping the clock
})[["elapsed"]]
t_anvil
```

## Runtime Comparison with Stan

As a sanity check, we compare the runtime of our single-chain {anvil} sampler against Stan via the \CRANpkg{rstan} package.
Stan's default sampler is NUTS, an adaptive variant of HMC that also tunes the step size and mass matrix during warmup.
To get an apples-to-apples comparison against our fixed-trajectory sampler, we switch Stan into static-HMC mode (`algorithm = "HMC"`), disable step-size and metric adaptation, and match the step size, integration time, and identity mass matrix we use in {anvil}.
Stan's static HMC then takes exactly the same number of leapfrog steps per iteration as ours (`int_time / stepsize = 16 / 0.2 = 80`).

Both frameworks pay a one-off compile cost before any sampling happens: Stan translates the model to C++ and compiles it with the system C++ compiler (tens of seconds), while {anvil} lowers the graph to StableHLO and hands it to the XLA JIT (a few seconds, paid once per unique static argument --- here, per choice of $n$).
To show how this overhead *amortises* as chains get longer, we vary $n$ over several orders of magnitude and measure the total wall-clock cost, both with a cold compilation cache and with the compiled executable already in hand.

```{r, stan-setup, message = FALSE, warning = FALSE}
stan_code <- "
data { real b; }
parameters { vector[2] theta; }
model {
  target += -square(theta[1]) / 200
            - square(theta[2] - b * square(theta[1]) + 100 * b) / 2;
}
"

stan_available <- tryCatch(
  {
    suppressMessages(library(rstan))
    rstan::rstan_options(auto_write = TRUE)
    t_stan_compile <- system.time(
      stan_model_banana <- rstan::stan_model(model_code = stan_code)
    )[["elapsed"]]
    # One warm-up sampling call so the first benchmarked call is not
    # paying for any additional first-time setup.
    suppressWarnings(rstan::sampling(
      stan_model_banana, data = list(b = b_param),
      chains = 1, iter = 100, warmup = 50,
      algorithm = "HMC",
      control = list(adapt_engaged = FALSE, stepsize = 0.2,
                     int_time = 16, metric = "unit_e"),
      refresh = 0, seed = 42L,
      save_warmup = FALSE, pars = "theta", include = TRUE
    ))
    TRUE
  },
  error = function(e) FALSE
)
```

```{r, bench-sweep, message = FALSE, warning = FALSE}
library(bench)
n_values <- c(100L, 1000L, 10000L, 50000L)

run_stan <- function(n) {
  suppressWarnings(rstan::sampling(
    stan_model_banana, data = list(b = b_param),
    chains = 1, iter = n + n_warmup, warmup = n_warmup,
    algorithm = "HMC",
    control = list(adapt_engaged = FALSE, stepsize = 0.2,
                   int_time = 16, metric = "unit_e"),
    refresh = 0, seed = 42L,
    save_warmup = FALSE, pars = "theta", include = TRUE
  ))
}

run_anvil <- function(n) {
  theta <- nv_array(matrix(0, nrow = 1L, ncol = 2L), dtype = "f64")
  rng_state <- nv_rng_state(42L)
  w <- hmc_sample_many(theta, rng_state, b_t, eps, L, n = n_warmup)
  out <- hmc_sample_many(w$theta, w$rng_state, b_t, eps, L, n = n)
  invisible(as_array(out$samples))  # force XLA synchronisation
}

# Pre-warm the JIT cache for n_warmup (used inside every run).
.dummy <- hmc_sample_many(
  nv_array(matrix(0, nrow = 1L, ncol = 2L), dtype = "f64"),
  nv_rng_state(1L), b_t, eps, L, n = n_warmup
)

# For each n we use bench::mark with four iterations: the first pays the
# XLA JIT compile for that n (reported as "cold"); the best of the remaining
# three is the steady-state cost (reported as "warm"). Stan is measured
# with the same best-of-three, against a model that was already compiled.
bench_row <- function(n) {
  anvil_times <- as.numeric(bench::mark(
    run_anvil(n),
    iterations = 4,
    check = FALSE, memory = FALSE, filter_gc = FALSE
  )$time[[1L]])
  stan_s <- if (stan_available) {
    min(as.numeric(bench::mark(
      run_stan(n),
      iterations = 3,
      check = FALSE, memory = FALSE, filter_gc = FALSE
    )$time[[1L]]))
  } else NA_real_
  data.frame(
    n = n,
    anvil_cold_s  = anvil_times[[1L]],
    anvil_warm_s  = min(anvil_times[-1L]),
    stan_sample_s = stan_s
  )
}
bench <- do.call(rbind, lapply(n_values, bench_row))
```

```{r, bench-table, echo = FALSE}
if (stan_available) {
  tab <- data.frame(
    n = bench$n,
    `anvil (with JIT compile)` = round(bench$anvil_cold_s, 3),
    `anvil (compiled)` = round(bench$anvil_warm_s, 3),
    `Stan sampling` = round(bench$stan_sample_s, 3),
    `Stan + model compile` = round(t_stan_compile + bench$stan_sample_s, 1),
    check.names = FALSE
  )
} else {
  tab <- data.frame(
    n = bench$n,
    `anvil (with JIT compile)` = round(bench$anvil_cold_s, 3),
    `anvil (compiled)` = round(bench$anvil_warm_s, 3),
    check.names = FALSE
  )
}
tab
```

A few observations:

- **Stan's model compile dominates short runs.** Translating and compiling the Stan model takes `r if (exists("t_stan_compile")) round(t_stan_compile, 1) else "~50"` seconds in our setup.
  Users only pay this once per R session (or never, if the compiled model is cached on disk via `auto_write = TRUE`), but for a fresh run it swamps any per-sample difference until chains are very long.
- **{anvil}'s JIT compile is much cheaper.** The graph-to-StableHLO translation and XLA compilation take roughly a second for this target, and the cost is paid once per unique value of the static argument `n`.
  In practice, scripts choose a single $n$ and amortise even this small cost.
- **After both compiles are paid, per-sample rates are in the same order of magnitude.**
  Stan and {anvil} each perform 80 leapfrog steps plus 81 gradient evaluations per iteration, and for this tiny two-dimensional target the work per step is small enough that both samplers are throughput-bound on similar things.
- **Stan still computes more per iteration than {anvil} does.**
  Even with `save_warmup = FALSE` and `pars = "theta"` telling Stan not to *return* the extra output, the sampler still computes per-iteration diagnostics (acceptance probability, energy, divergent-transition flag, tree statistics) and feeds them through its logging / output pipeline as a first-class part of each transition.
  Our {anvil} sampler does only the leapfrog and writes $\theta$. This inherent asymmetry means the comparison measures "total wall-clock to produce $n$ usable $\theta$ draws", not the pure leapfrog cost, and mildly favours {anvil}; stripping Stan's diagnostics would require modifying Stan itself.

Both samplers run **single-threaded** in this comparison, pinned explicitly via the environment variables we set at the top of the vignette.
For Stan, `chains = 1` already implies one sampling thread; the banana model contains no `reduce_sum(...)`, so within-chain parallelism cannot be used either.
For {anvil}, `PJRT_NPROC=1` caps XLA's intra-op thread pool and `--xla_cpu_multi_thread_eigen=false` disables Eigen's within-op threading in generated code.
The script `inst/verify-single-thread.R` confirms this by spawning a child R process with these flags, running a parallelism-hungry XLA-CPU workload, and sampling its CPU usage directly from `ps`; it can be re-run on any host to verify that the flags actually take effect on that machine.
Multi-core benefits for either framework appear only with many independent chains (run in parallel threads in Stan, or run as batched tensor operations in {anvil}, as we do next).

## Running Many Chains in Parallel

Even with the whole sampling loop inside a single JIT call, a chain of a two-dimensional target does only a tiny amount of work per leapfrog step.
On a GPU this is far below the level of parallelism the hardware can exploit.
The natural way to drive more work through the device is to run $K$ chains simultaneously, stacked as a $(K, 2)$ array.

Because our sampler was already batched, nothing about the kernel needs to change: we just hand it a $(K, 2)$ starting tensor.
The same `hmc_sample` and `hmc_sample_many` functions from above take it as-is --- no recompilation of source, only a recompilation of the XLA executable for the new input shape.

```{r}
K_chains <- 64L
n_warmup_b <- 200L
n_samples_b <- 200L

theta <- nv_array(matrix(0, nrow = K_chains, ncol = 2L), dtype = "f64")
rng_state <- nv_rng_state(42L)

for (i in seq_len(n_warmup_b)) {
  res <- hmc_sample(theta, rng_state, b_t, eps, L)
  theta <- res$theta
  rng_state <- res$rng_state
}

thetas <- vector("list", n_samples_b)
t_batch <- system.time({
  for (i in seq_len(n_samples_b)) {
    res <- hmc_sample(theta, rng_state, b_t, eps, L)
    theta <- res$theta
    rng_state <- res$rng_state
    thetas[[i]] <- theta
  }
  pooled <- do.call(rbind, lapply(thetas, as_array))
})[["elapsed"]]
```

Lifting the sampling loop into the graph works the same way for many chains as for one:

```{r}
theta <- nv_array(matrix(0, nrow = K_chains, ncol = 2L), dtype = "f64")
rng_state <- nv_rng_state(42L)
warm <- hmc_sample_many(theta, rng_state, b_t, eps, L, n = n_warmup_b)
theta <- warm$theta
rng_state <- warm$rng_state

t_batch_graph <- system.time({
  out <- hmc_sample_many(theta, rng_state, b_t, eps, L, n = n_samples_b)
  pooled <- matrix(as_array(out$samples), ncol = 2L)
})[["elapsed"]]
```

```{r, echo = FALSE}
data.frame(
  sampler = c(
    "anvil 1 chain (R-loop)",
    "anvil 1 chain (graph loop)",
    "anvil 64 chains (R-loop)",
    "anvil 64 chains (graph loop)"
  ),
  total_samples = c(n_samples, n_samples, K_chains * n_samples_b, K_chains * n_samples_b),
  seconds = round(c(NA_real_, t_anvil, t_batch, t_batch_graph), 3),
  samples_per_second = c(
    NA_integer_,
    round(n_samples / t_anvil),
    round(K_chains * n_samples_b / t_batch),
    round(K_chains * n_samples_b / t_batch_graph)
  ),
  row.names = NULL
)
```

Each time we push more of the work inside a single JIT-compiled call, the sample rate grows substantially:

1. **R loop with a per-iteration `jit` call**: every iteration pays the R $\to$ XLA dispatch cost, and that cost dominates for a small target.
2. **Sampling loop lifted into the graph**: a single call produces all samples; dispatch overhead is amortised across the full chain.
3. **Parallel chains**: the leapfrog integrator now operates on $(K, 2)$ tensors, so each gradient and update does $K$ times more arithmetic for the same dispatch cost.
4. **Both combined**: a single JIT call produces $n \cdot K$ samples, running the leapfrog as batched tensor operations, with no R round-trips and no dispatch overhead on the inner loop.

On a GPU the benefit of (3) and (4) grows further: the leapfrog integrator becomes a sequence of dense matrix operations of shape $(K, 2)$, which is exactly the regime for which XLA's GPU backend is designed.
The same compiled function runs on CPU or GPU without code changes --- only the backend chosen at jit-time differs.
