---
title: "JIT Deep Dive"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{JIT Deep Dive}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains what actually happens when you wrap a function with `jit()`.
Understanding this is what lets you avoid the classic pitfalls that come with tracing-based compilation.
If you haven't yet, read the [Get Started](anvl.html) vignette first.

We will use a single tiny `linear` function as the running example throughout, extending it as we go:

```{r}
library(anvl)
set.seed(42)

linear <- function(x, w, b) nv_add(nv_mul(x, w), b)
```

We spell the operations out as `nv_add()` / `nv_mul()` for clarity; `x * w + b` would be equivalent because the standard arithmetic operators are overloaded for `AnvlArray`s and dispatch to the same `nv_*` functions.

## How `jit()` works

In pseudo-R, `jit(f)` returns roughly this closure:

```r
jit <- function(f, static = character()) {
  cache <- hashtab()
  function(...) {
    # abstract representations of dynamic args, values of static args
    key <- input_signature(..., static = static)
    if (is.null(cache[[key]])) {
      # step 1: record prim_* calls into an AnvlGraph
      graph <- trace_fn(f, list(...), static = static)
      # step 2: lower to an XLA executable, store
      cache[[key]] <- compile(graph)
    }
    # step 3: run the cached executable on this call's inputs
    cache[[key]](...)
  }
}
```

Three things happen:

1. **Trace** the R code once with placeholder values (except for `static` arguments; c.f. next section), recording the sequence of primitive operations into an intermediate representation called an `AnvlGraph`.
1. **Compile** that graph to an XLA executable and cache it under a key derived from the inputs, which uses abstract types for `AnvlArray`s and the actual values for static inputs.
1. On subsequent calls with a cache hit, **skip tracing and compilation** and run the cached executable directly.

### The compilation cache

When a call's inputs match a cached entry, both tracing and compilation are skipped.
We can observe this directly by inspecting the size of the cache held inside the jitted function:

```{r}
cache_size <- function(f) environment(f)$cache$size

linear_jit <- jit(linear)
cache_size(linear_jit)   # 0: nothing cached yet

linear_jit(nv_scalar(2), nv_scalar(3), nv_scalar(1))
cache_size(linear_jit)   # 1: a new entry was added

# same shapes -> cache hit, size unchanged
linear_jit(nv_scalar(2), nv_scalar(3), nv_scalar(5))
cache_size(linear_jit)

# different shapes -> a second entry is added
linear_jit(
  nv_array(c(1, 2)),
  nv_array(c(3, 4)),
  nv_array(c(1, 1))
)
cache_size(linear_jit)
```

Each input to the function contributes to the cache key differently, depending on whether it is *dynamic* (an arrayish value, the default) or *static* (marked via the `static =` argument of `jit()`):

- **Dynamic inputs** contribute their *abstract value* -- the `nv_aval(dtype, shape, ambiguous)` triple.
  The concrete data inside the array is ignored, which is why calls with the same shape and dtype hit the same cache entry no matter what values they carry.
  During tracing (next section) these are replaced by type-only placeholders.
- **Static inputs** contribute their *exact R value*, compared with `identical()`.
  They stay as regular R values during the compilation, but their value is fixed for a compiled program.
  Two calls with `flag = TRUE` and `flag = FALSE` therefore land on different cache entries.

In the snippet above all three inputs are dynamic, so the first two calls share a key (three `f32[]` scalars) and hit the cache, while the third call presents three `f32[2]` vectors and forces a retrace.

You'll want `static =` when the body of the function needs to look at a concrete R value -- typically a flag, a small integer, or a string.
It's also the only way to do R-level input validation on a value: a dynamic input is just a shape/dtype placeholder during tracing, so a check like `stopifnot(abs(sum(p) - 1) < 1e-6)` -- verifying that `p` is a proper probability vector -- only works if `p` is static.

```{r}
linear_maybe <- function(x, w, b, use_bias) {
  if (use_bias) linear(x, w, b) else x * w
}
linear_maybe_jit <- jit(linear_maybe, static = "use_bias")

linear_maybe_jit(2, 3, 1, use_bias = TRUE)
linear_maybe_jit(2, 3, use_bias = FALSE)
```

Each call with a new static value forces a re-trace and re-compile, so static arguments cause more re-compiles than dynamic ones.
Only use them when you really need them.

The cache key also includes information about the input structure (if `AnvlArray`s are nested in lists) or the device to compile for, but we will ignore them as they are less relevant to understand for using `jit()`.

Two important notes to be aware of:

- Call `jit()` once on a function and reuse the result; calling `jit()` inside a loop creates a fresh cache on every iteration and defeats the point:
- For functions that will be called many times with consistent shapes, compilation is a one-time cost.
  For computations on shapes that won't recur, the compile time may dominate -- see [Padding Inputs to Avoid Recompilation](efficiency.html#padding-inputs-to-avoid-recompilation) in the efficiency vignette for one way to keep the cache small.

## Tracing

*Tracing* is what we call running your R function once with special values and recording every primitive operation it performs into an `AnvlGraph`.
For the purposes of this vignette you can think of tracing and the subsequent XLA compilation as a single phase that runs once per cache miss.
Note that this is different from R's `trace()` function, which lets you insert code into functions.
Tracing is how we leave R behind: the graph is a self-contained, language-agnostic description of the computation that anvl can hand off to the XLA compiler.
It's much simpler than R code, which also enables differentiating it.
The graph is the evaluation trace, and most of what follows in this vignette is about understanding which type of programs can properly be captured this way (taking into account the caching mechanism discussed above).

`trace_fn()` is the lower-level building block that `jit()` uses internally.
Although you will probably never need to use this function directly, we will use it to show what's happening under the hood.
Tracing only needs to know the *type* of each dynamic input -- its shape, dtype, and ambiguity flag -- not its concrete value (except for `static` inputs).
We describe such a type-only placeholder with `nv_aval()` (short for *abstract value*):

```{r}
nv_aval("f32", integer())                    # an f32 scalar
nv_aval("f32", c(2, 3))                      # a 2x3 f32 matrix
nv_aval("f32", integer(), ambiguous = TRUE)  # ambiguous f32 (printed as `f32?`)
```

An abstract value carries no data; it just stands in for "any array of this shape, dtype, and ambiguity" during the trace.
Concrete `AnvlArray`s also work as `trace_fn` arguments -- their shape, dtype, and ambiguity are read off and the data is ignored -- but `nv_aval()` makes it explicit that we only care about the type.
(See `vignette("type-promotion")` for a refresher on what ambiguity means.)

Let's trace `linear` with a length-3 vector for `x` and scalars for `w` and `b`:

```{r}
f32_scalar <- nv_aval("f32", integer())
f32_vec3   <- nv_aval("f32", 3)
trace_fn(linear, args = list(x = f32_vec3, w = f32_scalar, b = f32_scalar))
```

A few things to notice:

- The R-level *expression structure* is flattened. The nested call `nv_add(nv_mul(x, w), b)` becomes a sequence of separately-named intermediate values; each operation produces a new graph value, so the IR is single-assignment and purely functional.
- The shape and dtype of every value are part of the graph. You can see the inputs come in with their abstract types attached, and the types propagate through the operations.
- The function call boundary is gone: if `linear` had called another R function, that function's body would have been inlined into the same graph.
- A single `nv_*` call can lower to *several* primitives. The `nv_mul(x, w)` here doesn't appear as one `mul` op -- the scalar `w` first goes through a broadcast primitive that lifts it to the shape of `x`, and only then is the actual element-wise `mul` recorded. Same for `nv_add(..., b)`.

Crucially, only `prim_*` calls (and the `nv_*` API functions or overloaded operators that delegate to them) get recorded.
Any other R code in the body might influence what the evaluation trace is, but is not present in the traced graph itself.

### The Tracing Contract

Tracing only produces correct results for *pure* functions -- functions whose output depends on their arguments and nothing else, and that have no side effects.
This means, there should be:

- **No hidden inputs.** If the function reads anything outside its arguments -- a closed-over R value, the random-number generator, the system clock, a global option -- the trace silently bakes in whatever it observed during the single tracing call, and subsequent calls don't re-observe.
- **No side effects.** Any R-level mutation or I/O (printing, `<<-`, mutating an enclosing environment, writing to a file) fires only while the trace is being built; the cached executable contains only `prim_*` operations, so on subsequent calls the side effect simply doesn't happen.

This contract is a consequence of how the caching works: re-tracing on every call would defeat the point of compilation, so the cache key captures only the abstract type of each dynamic input and the exact value of each static input.
Concretely, the function's execution path -- the specific sequence of primitive calls it performs -- must depend only on:

1. The *abstract* representation (shape, dtype, ambiguity) of each dynamic input, and
2. The *value* of each static input.

The next subsections each show what tracing does to particular R code and where its behavior might be surprising.

### R loops are unrolled

Tracing runs your R code and records primitive operations in the graph.
Because `for` is not an anvil primitive, it will be executed as usual and all the primitive calls encountered will be recorded in the graph.
Here we apply the `linear` function `n` times.

```{r}
linear_repeated <- function(x, w, b, n) {
  for (i in seq_len(n)) x <- linear(x, w, b)
  x
}
trace_fn(linear_repeated, args = list(x = f32_scalar, w = f32_vec3, b = f32_vec3, n = 2L))
```

The graph contains a single `broadcast_in_dim` (lifting the scalar `x` to `f32[3]`) followed by two `mul`/`add` pairs in sequence -- not a loop construct.
The broadcast happens only once because by the second iteration `x` is already an `f32[3]` and matches `w` and `b` directly.
The compiled executable will contain those five operations laid out one after another.
For a different value of `n`, the loop would get unrolled for this specific iteration number.
Long loops also lead to long compile times and large executables.
You can instead use `nv_while()`, which records a single higher-order primitive call in the graph regardless of how many iterations the loop runs.

### R `if` statements pick one branch

Tracing runs only the `if`-branch that the condition selects at trace time.
One common scenario is where the branch depends on a static input flag:

```{r}
linear_maybe <- function(x, w, b, use_bias) {
  if (use_bias) linear(x, w, b) else x * w
}
trace_fn(
  linear_maybe,
  args = list(x = f32_scalar, w = f32_scalar, b = f32_scalar, use_bias = TRUE)
)
```

The graph contains one `mul` and one `add` operation, but no conditional.
Tracing with `use_bias = FALSE` would produce a different graph.

Where things go wrong is if the evaluation trace depends on something that does not influence the cache key, such as a value from the enclosing environment:

```{r}
threshold <- 0.5
h <- function(x) {
  if (threshold > 0.5) x * 2 else x + 1
}
trace_fn(h, args = list(x = f32_scalar))
```

The trace itself runs correctly, but it becomes a problem in combination with `jit()`'s caching mechanism.
The closed-over `threshold` does not influence the cache key, so subsequent calls with the same dynamic input type would reuse the executable produced when `threshold` was `0.5`, no matter how `threshold` is at call time.
The only fix is to not let the graph depend on values outside the function's signature -- in this case, make `threshold` an explicit static argument.

### Closed-over values become constants

We just saw an extreme version of this in the `if` section: a closed-over R variable picked the *branch* that ended up in the graph.
The same dynamic applies to closed-over values used as plain operands -- their value at trace time is read once and baked into the graph.

Here we close over a default bias instead of taking it as an argument:

```{r}
default_b <- 5
linear_default_b <- function(x, w) linear(x, w, default_b)
trace_fn(linear_default_b, args = list(x = f32_scalar, w = f32_scalar))
```

The graph contains `add(%1, 5:f32?)` -- the value `5` is hard-wired into the program.
Changing `default_b` afterwards has no effect on the graph (and, once the graph is compiled, no effect on the executable either).

### Side effects only fire during tracing

The *Tracing Contract* above already noted that a jitted function must be pure. This subsection makes the consequence concrete: **R-level side effects only have an effect while the graph is being built**, not on subsequent calls.

A common R pattern for stateful objects is to wrap them in an environment, since environments give you reference semantics:

```{r}
new_model <- function(beta) {
  e <- new.env()
  e$beta <- beta
  e$grad_step <- function(beta_grad, lr) {
    e$beta <- e$beta - beta_grad * lr
    e$beta
  }
  e
}
```

One would expect each call to `model$grad_step()` to nudge `model$beta` one step further along the gradient.
But wrapping `grad_step` with `jit()` breaks the function on two levels:

```{r}
model <- new_model(nv_array(c(0, 0, 0), dtype = "f32"))
grad_step_jit <- jit(model$grad_step)
g <- nv_array(c(1, 1, 1), dtype = "f32")

grad_step_jit(g, 0.1)   # expected c(-0.1, -0.1, -0.1)
grad_step_jit(g, 0.1)   # expected c(-0.2, -0.2, -0.2) -- but identical to call 1
grad_step_jit(g, 0.1)   # expected c(-0.3, -0.3, -0.3) -- but identical to call 1

class(model$beta)       # not even an AnvlArray any more
```

Two things went wrong, both for the same reason: the only thing `jit()` records into the graph is `prim_*` calls.

- The mutation `e$beta <- ...` is a plain R assignment, not a `prim_*` call, so it doesn't get recorded into the `AnvlGraph`.
  It runs once, during tracing -- but the value being assigned is anvl's internal trace placeholder (a `GraphBox`), not a real array.
  So after the first call, `model$beta` is *broken*: it holds a leaked tracer instead of the array you expected.
- On every subsequent call only the cached XLA executable runs -- never the R body, so the `e$beta <- ...` line will not be executed again.
  And because `e$beta` was a closed-over R value at trace time, its initial value (`c(0, 0, 0)`) is baked into the graph as a literal (per *Closed-over values become constants* above), so every call returns the same `-0.1`.

This is why `jit()`-compiled functions must be *pure*: their output depends only on their arguments, and any state updates have to happen at the call site, not inside the function:

```{r}
grad_step <- jit(function(beta, beta_grad, lr) beta - beta_grad * lr)

beta <- nv_array(c(0, 0, 0), dtype = "f32")
beta <- grad_step(beta, g, 0.1)
beta <- grad_step(beta, g, 0.1)
beta <- grad_step(beta, g, 0.1)
beta
```

## Other `jit()` arguments

### Donating inputs

By default, a compiled executable treats its inputs as read-only: the R-visible input arrays remain valid after the call, and XLA has to allocate fresh memory for any output of matching shape.
For long training loops over large parameters, this means every step allocates a new parameter buffer and leaves the previous one for the garbage collector.

Via the `donate` argument of `jit()`, you can tell XLA that an input will not be used after the call, so it is free to reuse the input array's memory for an output.

```{r}
step <- jit(function(w, g) w - 0.1 * g, donate = "w")

w <- nv_array(c(1, 2, 3), dtype = "f32")
g <- nv_array(c(0.1, 0.1, 0.1), dtype = "f32")

w <- step(w, g)
w
```

The compiled executable now consumes `w`'s buffer as part of the call.
The caller must not reuse an array that was donated to a function, otherwise an error is thrown:

```{r, error = TRUE, eval = Sys.getenv("PJRT_PLATFORM") == "cpu"}
w_old <- nv_array(c(1, 2, 3), dtype = "f32")
w_new <- step(w_old, g)
w_old     # the old buffer has been donated
```

### Device placement

Every `AnvlArray` lives on a concrete device (CPU, a specific GPU, etc.), and a compiled executable is itself bound to a specific device -- so all inputs to one call must live on that same device.
How the device is chosen for a `jit()`-compiled call depends on how you called `jit()`.
The two most relevant options are:

- **Inferred from inputs (default).** If you don't pass `device =`, anvl looks at the devices of the array inputs at call time, requires them to agree, and compiles for that device.
  If there are no array inputs, it falls back to the `default_device()`.
- **Pinned at `jit()` time.** Passing a concrete device -- e.g. `jit(f, device = "cpu:0")` or `jit(f, device = nv_device("cuda:0"))` -- forces every call to run on that device.
  Inputs living on a different device are copied over automatically.

See the documentation of the `device` arg in `help(jit)` for more information.

```{r}
add <- jit(function(x, y) x + y)
add_cpu <- jit(function(x, y) x + y, device = "cpu")

a <- nv_array(c(1, 2, 3), dtype = "f32")
b <- nv_array(c(4, 5, 6), dtype = "f32")

device(add(a, b))                            # inferred from inputs
device(add_cpu(a, b))                        # pinned
```

Because the device is part of the cache key (see *The compilation cache* above), a single `jit()`ted function can hold compiled binaries for several devices at once, unless `device` was specified explicitly during `jit()`.
