---
title: "Metropolis-Hastings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metropolis-Hastings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we show how to implement a random-walk Metropolis-Hastings
sampler using {anvil}.
Metropolis-Hastings is a foundational MCMC algorithm that generates samples from a target distribution by proposing moves and accepting or rejecting them based on the density ratio.

## The Banana Distribution

We sample from a "banana-shaped" distribution (`r anvil:::cite_bib("haario_1999")`).
This 2-dimensional distribution consists of two gaussian densities, where the second is dependent on the first:

$$p(\theta_1, \theta_2) = \underbrace{\frac{1}{\sqrt{2\pi\sigma^2}} \exp\!\left(-\frac{\theta_1^2}{2\sigma^2}\right)}_{p(\theta_1)} \;\cdot\; \underbrace{\frac{1}{\sqrt{2\pi}} \exp\!\left(-\frac{\bigl(\theta_2 - b(\theta_1^2 - \sigma^2)\bigr)^2}{2}\right)}_{p(\theta_2 \mid \theta_1)}$$

While it is trivial to sample from this distribution, it serves as a useful test case for our own Metropolis-Hastings sampler.

The distribution has known analytical moments:

- $\mathbb{E}[\theta_1] = 0$, $\text{Var}[\theta_1] = \sigma^2$
- $\mathbb{E}[\theta_2] = 0$, $\text{Var}[\theta_2] = 2 b^2 \sigma^4 + 1$

The mean of $\theta_2$ follows from
$\mathbb{E}[\theta_2] = b(\mathbb{E}[\theta_1^2] - \sigma^2) = 0$,
and its variance from the law of total variance:
$\text{Var}[\theta_2] = b^2 \text{Var}[\theta_1^2] + 1 = 2b^2\sigma^4 + 1$.
These closed-form moments make this distribution a useful test case,
since we can verify that our sampler recovers the correct values.

We use $\sigma^2 = 1$ and $b = 1$, giving
$\text{SD}[\theta_1] = 1$ and $\text{SD}[\theta_2] = \sqrt{3} \approx 1.73$:

```{r}
library(anvil)
set.seed(123)

sigma2 <- 1
b_param <- 1
```

```{r, echo = FALSE, fig.width = 7, fig.height = 5}
log_p <- function(x1, x2) {
  -x1^2 / (2 * sigma2) - (x2 - b_param * (x1^2 - sigma2))^2 / 2
}

x1_grid <- seq(-4, 4, length.out = 200)
x2_grid <- seq(-4, 8, length.out = 200)
z <- outer(x1_grid, x2_grid, log_p)

par(mar = c(4, 4, 2, 1))
contour(
  x1_grid, x2_grid, exp(z), nlevels = 15,
  xlab = expression(theta[1]), ylab = expression(theta[2]),
  main = "Banana Distribution", col = "grey40"
)
```

## Log-Density

The corresponding log-density (up to a constant) is:

$$\log p(\theta_1, \theta_2) = -\frac{\theta_1^2}{2\sigma^2} - \frac{\bigl(\theta_2 - b(\theta_1^2 - \sigma^2)\bigr)^2}{2}$$

We implement this in {anvil} and convert the parameters to `AnvilTensor`s:

```{r}
sigma2_t <- nv_scalar(sigma2, dtype = "f32")
b_t <- nv_scalar(b_param, dtype = "f32")

log_density <- function(theta, sigma2, b) {
  theta1 <- theta[1]
  theta2 <- theta[2]
  -theta1^2 / (2 * sigma2) - (theta2 - b * (theta1^2 - sigma2))^2 / 2
}
```

## Metropolis-Hastings Algorithm

The random-walk Metropolis-Hastings algorithm proceeds as follows:

1. Propose $\theta' = \theta + \varepsilon$, where $\varepsilon \sim \mathcal{N}(0, s^2 I)$
2. Compute the log acceptance ratio $\log \alpha = \log p(\theta') - \log p(\theta)$
3. Accept $\theta'$ with probability $\min(1, \alpha)$

The accept/reject decision uses `nv_ifelse()`.

```{r}
mh_step <- function(theta, rng_state, sigma2, b, proposal_sd) {
  rng_out <- nv_rnorm(shape(theta), rng_state, dtype = "f32")
  rng_state <- rng_out[[1L]]
  proposal <- theta + proposal_sd * rng_out[[2L]]

  log_current <- log_density(theta, sigma2, b)
  log_proposed <- log_density(proposal, sigma2, b)
  log_accept <- log_proposed - log_current

  rng_out <- nv_runif(1L, rng_state, dtype = "f32")
  rng_state <- rng_out[[1L]]
  u <- rng_out[[2L]]

  accept <- log(u) < log_accept
  new_theta <- nv_ifelse(
    nv_broadcast_to(accept, shape_abstract(theta)),
    proposal,
    theta
  )

  list(theta = new_theta, rng_state = rng_state)
}
```

We use `nv_while()` to run multiple thinned steps in a single JIT-compiled
call, avoiding R loop overhead:

```{r}
mh_sample <- jit(function(theta, rng_state, sigma2, b, proposal_sd, thin) {
  result <- nv_while(
    init = list(theta = theta, rng_state = rng_state, i = nv_scalar(0L)),
    cond = \(theta, rng_state, i) i < thin,
    body = \(theta, rng_state, i) {
      step <- mh_step(theta, rng_state, sigma2, b, proposal_sd)
      list(theta = step$theta, rng_state = step$rng_state, i = i + 1L)
    }
  )[1:2]
})
```

## Running the Sampler

We run the sampler from R, collecting samples after a warmup period.
The first call triggers JIT compilation; subsequent calls reuse the compiled
code. Because random-walk proposals on a curved distribution produce highly
correlated chains, we thin the output by keeping every 100th sample.

```{r}
n_samples <- 20000L
n_warmup <- 500L
thin <- nv_scalar(100L)

theta <- nv_tensor(c(0, 0), dtype = "f32")
rng_state <- nv_rng_state(seed = 42L)
proposal_sd <- nv_scalar(0.5, dtype = "f32")

result <- mh_sample(theta, rng_state, sigma2_t, b_t, proposal_sd, thin)
theta <- result$theta
rng_state <- result$rng_state

for (i in seq_len(n_warmup)) {
  result <- mh_sample(theta, rng_state, sigma2_t, b_t, proposal_sd, thin)
  theta <- result$theta
  rng_state <- result$rng_state
}

samples <- matrix(NA_real_, nrow = n_samples, ncol = 2L)
colnames(samples) <- c("theta1", "theta2")

for (i in seq_len(n_samples)) {
  result <- mh_sample(theta, rng_state, sigma2_t, b_t, proposal_sd, thin)
  theta <- result$theta
  rng_state <- result$rng_state
  samples[i, ] <- as.vector(as_array(theta))
}
```

## Results

```{r}
sample_mean <- colMeans(samples)
sample_sd <- apply(samples, 2, sd)

cat("Sample means:\n")
print(round(sample_mean, 3))
cat("\nSample standard deviations:\n")
print(round(sample_sd, 3))
```

## Verification

The banana distribution has known analytical moments.
Since $\theta_1 \sim \mathcal{N}(0, \sigma^2)$ and
$\theta_2 \mid \theta_1 \sim \mathcal{N}(b(\theta_1^2 - \sigma^2), 1)$:

- $\mathbb{E}[\theta_1] = 0$, $\text{SD}[\theta_1] = \sigma$
- $\mathbb{E}[\theta_2] = 0$, $\text{SD}[\theta_2] = \sqrt{2 b^2 \sigma^4 + 1}$

```{r, echo = FALSE}
true_mean <- c(0, 0)
true_sd <- c(sqrt(sigma2), sqrt(2 * b_param^2 * sigma2^2 + 1))

comparison <- data.frame(
  Parameter = c("theta1", "theta2"),
  MH_Mean = sample_mean,
  True_Mean = true_mean,
  MH_SD = sample_sd,
  True_SD = true_sd,
  row.names = NULL
)
comparison
```

The Metropolis-Hastings estimates closely match the analytical moments,
confirming the correctness of our implementation.

We can visualize the samples against the true density contours:

```{r, echo = FALSE, fig.width = 7, fig.height = 5}
par(mar = c(4, 4, 2, 1))
contour(
  x1_grid, x2_grid, exp(z), nlevels = 15,
  xlab = expression(theta[1]), ylab = expression(theta[2]),
  main = "MH Samples vs True Density", col = "grey60"
)
points(
  samples[, 1], samples[, 2],
  pch = 16, cex = 0.3, col = adjustcolor("steelblue", 0.4)
)
```

And overlay the analytical marginal densities on histograms of the samples.
The marginal of $\theta_1$ is Gaussian; the marginal of $\theta_2$ is
obtained by numerically integrating out $\theta_1$:

```{r, echo = FALSE, fig.width = 7, fig.height = 4}
marginal_theta2 <- Vectorize(function(t2) {
  integrate(function(t1) {
    dnorm(t2, mean = b_param * (t1^2 - sigma2), sd = 1) *
      dnorm(t1, mean = 0, sd = sqrt(sigma2))
  }, lower = -6, upper = 6)$value
})

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

hist(
  samples[, 1], breaks = 40, probability = TRUE,
  main = expression(theta[1]),
  xlab = expression(theta[1]), col = "lightblue", border = "white"
)
curve(
  dnorm(x, mean = 0, sd = sqrt(sigma2)),
  add = TRUE, col = "red", lwd = 2
)
legend("topright", "Analytical", col = "red", lwd = 2, bty = "n", cex = 0.8)

hist(
  samples[, 2], breaks = 40, probability = TRUE,
  main = expression(theta[2]),
  xlab = expression(theta[2]), col = "lightblue", border = "white"
)
curve(marginal_theta2, add = TRUE, col = "red", lwd = 2, n = 200)
legend("topright", "Analytical", col = "red", lwd = 2, bty = "n", cex = 0.8)
```
