---
title: "Adding a Primitive"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a Primitive}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This guide explains how to implement a new primitive.
It will primarily focus on *how* to do this.
See the [internals vignette](internals.html) for more information on how primitives work.

In general, there are two main reasons to add a new primitive:

1. The operation is not expressible using existing primitives.
2. The operation (or, for example, its derivative) would benefit from a dedicated implementation.

In order to be able to add a new primitive, it needs to be expressible in the [{stablehlo}](https://github.com/r-xla/stablehlo) package.
There are various sceanrios:

1. The operation is expressible in the StableHLO language and:
   1. All required operations are already implemented in the {stablehlo} R package.

      $\rightarrow$ This is the simplest caseand the one we assume in this guide.
   2. One or more required operations are not already implemented in the {stablehlo} R package.

      $\rightarrow$ You first need to implement the missing operations in the {stablehlo} R package.
      See [this issue](https://github.com/r-xla/stablehlo/issues/6) for which operations are missing and the [StableHLO specification](https://openxla.org/stablehlo/spec) for which are available.
2. The operation is not available/cannot be efficiently expressed in StableHLO, because:
   1. It requires shape dynamism (output shape depends on data, such as using boolean values for subsetting):

      $\rightarrow$ This is currently not possible, but we hope we can add support for this in the future, e.g. via a second, dynamic Fortran backend.

   2. It cannot be expressed or can only expressed inefficiently:

      $\rightarrow$ You can implement a stableHLO custom call, see the custom print operation in [pjrt](https://github.com/r-xla/pjrt).
      This is currently not well documented, so you need to dig into the source code.

## Adding a Primitive: Practical Example

Let's add a new primitive step by step.
We'll implement `prim_repeat_along` -- a primitive that repeats an array multiple times along a specified dimension.

For example, repeating `c(1, 2, 3)` twice along dimension 1 gives `c(1, 2, 3, 1, 2, 3)`.

This primitive has a *dynamic* input (an array) and two *static* parameters (how many times to repeat and which dimension).

### Step 1: Define the Primitive

Primitives are created with `new_primitive()`: it builds the `AnvilPrimitive` metadata object that holds the rules, wraps the body with `jit()`, attaches the metadata, and registers the result into the exported `prim` environment. The returned callable becomes the primitive.

```{r}
library(anvil)
prim_repeat_along <- new_primitive(
  "repeat_along",
  function(operand, times, dim) {
    # type of operand is checked by graph_desc_add()
    infer_fn <- function(operand, times, dim) {
      if (!checkmate::test_integerish(dim, lower = 1, upper = ndims(operand), len = 1L)) {
        cli::cli_abort("{.arg dim} must be between 1 and {ndims(operand)}, but is {.val dim}")
      }
      if (!checkmate::test_integerish(times, lower = 1, len = 1L)) {
        cli_abort("times must be a positive integer, but is {times}")
      }
      new_shape <- shape(operand)
      new_shape[dim] <- new_shape[dim] * times
      list(AbstractArray(
        dtype = dtype(operand),
        shape = Shape(new_shape),
        ambiguous = operand$ambiguous
      ))
    }

    graph_desc_add(
      "repeat_along",             # primitive name — looked up in `prim`
      list(operand = operand),    # Dynamic inputs (arrays)
      params = list(              # Static parameters
        times = times,
        dim = dim
      ),
      infer_fn = infer_fn
    )[[1L]]  # Extract single output from list
  },
  static = c("times", "dim")
)
```

The primitive is now callable directly (`prim_repeat_along(x, times, dim)`) and also accessible via `prim$repeat_along`.

Key points:

- Pass the primitive **name string** to `graph_desc_add()`; it looks up the primitive in the `prim` registry.
- The inference function receives abstract arrays (types, not values) for dynamic inputs, and actual values for static parameters. It must verify that the arguments to the function are valid, and should throw clear error messages — {anvil} programs are otherwise hard to debug.
- `graph_desc_add()` returns a list of outputs; use `[[1L]]` for single-output primitives.
- Propagate the `ambiguous` flag from inputs to outputs, see [type promotion](type-promotion.html) for what this means.

Every argument that is *not* a dynamic array must be listed in `static`. `new_primitive` calls `jit()` with `backend = "auto"` internally, which defers the choice of backend to call time.

If the primitive is a wrapper around a stablehlo operation, it is possible to use the corresponding inference function from the stablehlo package (such as `stablehlo::infer_types_concatenate`).
When doing so, you need to:

1. Convert the abstract arrays to stablehlo `ValueType`s using `at2vt()`.
2. Call the stablehlo inference function and obtain a list of `ValueType`s.
3. Convert the `ValueType`s back to abstract arrays using `vt2at()`.
4. Set the `ambiguous` flag of the output depending on the inputs (`ambiguity` is strictly an {anvil} concept, not a stablehlo concept).

#### Special case: primitives with 0 dynamic inputs

Array *constructors* (e.g. `prim$fill`, `prim$iota`) have no dynamic array inputs -- every argument is static. Two things change in this case:

1. **All formals must be listed in `static`**. Otherwise {anvil} would try to interpret a scalar argument like `value` or `shape` as an `AnvilArray` input.
2. **`backend = "auto"` cannot infer a device from inputs**, because there are no array inputs. Instead, add a `device` formal to the function and pass `device = device_arg("device")` to `new_primitive()`. At call time, the user-supplied device is read from that argument and used to determine both the backend (via [`backend()`] dispatch) and the compilation device.

For example, the real definition of `prim$fill` looks like this:

```r
prim_fill <- new_primitive(
  "fill",
  function(value, shape, dtype, ambiguous = FALSE, device = NULL) {
    # ... graph_desc_add("fill", ...) ...
  },
  static = 1:5,
  device = device_arg("device")
)
```

See [`device_arg()`] for the detailed semantics.

### Step 3: Add the StableHLO Rule

The StableHLO rule defines how to lower this primitive to actual operations, which is used in the `stablehlo()` lowering pass.
We implement `repeat_along` using concatenation:

```{r}
prim_repeat_along[["stablehlo"]] <- function(operand, times, dim) {
  operands <- rep(list(operand), times)
  list(rlang::exec(stablehlo::hlo_concatenate, !!!operands, dimension = dim - 1L))
}
```

The rule receives:

- Dynamic inputs as `stablehlo::FuncValue`s.
- Static parameters as their R values.

It must return a list of `stablehlo::FuncValue`s, even if there is only one output.

**Important**: StableHLO uses 0-based indexing, while {anvil} uses R's 1-based indexing.
Always convert dimension indices by subtracting 1.
Also note that in `graph_desc_add()` we are converting the error messages from stablehlo to our 1-based indexing, so you do not have to worry about that here.

### Step 4: Add the Reverse Rule

If the operation should support automatic differentiation, add a reverse rule.
The idea here is the following, where we assume the input `operand` has shape `(s_1, ..., s_n)`, which means that the output (and therefore it's gradient) has shape `(s_1, ..., s_{dim-1}, s_dim * times, s_{dim+1}, ..., s_n)`.

1. Reshape the gradient to `(s_1, ..., s_{dim-1}, s_dim, times, s_{dim+1}, ..., s_n)`.
2. Sum over the `times` dimension and drop the `times` dimension.

```{r}
prim_repeat_along[["reverse"]] <- function(inputs, outputs, grads, dim, times, .required) {
  if (!.required[[1L]]) {
    return(list(NULL))
  }

  grad <- grads[[1L]]
  operand <- inputs[[1L]]

  old_shape <- shape(operand)
  grad_shape <- shape(grad)

  new_shape <- grad_shape
  new_shape[dim] <- old_shape[dim]
  new_shape <- append(new_shape, times, after = dim - 1L)

  grad_reshaped <- prim_reshape(grad, new_shape)
  grad_summed <- prim_reduce_sum(grad_reshaped, dims = dim, drop = TRUE)
  list(grad_summed)
}
```

The reverse rule receives:

- `inputs`: Input `GraphValue`s from the forward pass
- `outputs`: Output `GraphValue`s from the forward pass
- `grads`: Gradients flowing back from downstream (one per output)
- Static parameters by name (here: `dim`, `times`)
- `.required`: Logical vector indicating which input gradients are needed

It returns a list with one gradient per input (or `NULL` if not required).

### Step 5: Verify the Registration

`new_primitive()` registers the primitive into the exported `prim` environment as a side effect, so no separate registration step is needed:

```{r}
prim$repeat_along
```

### Step 6: Add an `nv_` API Function

In {anvil}, we also offer convenience wrappers around the primitives.
An example is `prim$add` vs `nv_add`, where the latter calls into the former after optionally broadcasting (scalar) inputs:

```{r, error = TRUE}
nv_add(1L, nv_array(2:3))
prim$add(1L, nv_array(2:3))
```

In our case, no such convenience is needed and the functionality is not too low-level (for it to be generally useful), so we can just reassign the `prim_*` function to an `nv_*` function:

```{r}
nv_repeat_along <- prim_repeat_along
```

Note that in the `nv_*` wrapper function, you can only access certain properties of the input arrayish values via:

- `shape_abstract()`
- `ndims_abstract()`
- `dtype_abstract()`
- `ambiguous_abstract()`

If you, for example, use `shape()` instead of `shape_abstract()`, your function won't work with R literals.
I.e., `<extract>_abstract()` first converts the input to an `AbstractArray` (if possible) and then extracts the property.

### Using Your Primitive

You can now use the primitive, both eagerly and inside a larger JIT-compiled function:

```{r}
x <- nv_array(c(1, 2, 3), shape = c(3, 1))
# Eager call -- works because prim_repeat_along is itself jit()ted.
prim_repeat_along(x, times = 2L, dim = 2L)
# Traced into the outer graph when composed with another jit()ted function.
jit(function(x) prim_repeat_along(x, times = 2L, dim = 2L))(x)
```

And compute gradients through it.

```{r}
f <- function(x) {
  repeated <- prim_repeat_along(x, times = 2L, dim = 2L)
  sum(repeated)
}

grad_f <- jit(gradient(f))
grad_f(x)[[1L]]
```

## Contributing to the Package

If you want to contribute your primitive to {anvil}, there are some additional things to be aware of.

### File Organization

- **`R/primitives.R`**: Define `AnvilPrimitive` object and `prim_*` function
- **`R/rules-stablehlo.R`**: Add the StableHLO lowering rule
- **`R/rules-reverse.R`**: Add the reverse rule (if differentiable)
- **`R/api.R`**: Add the `nv_*` wrapper function (or possibly in another **api** file).

### Testing

Tests can go in two places:

1. **`inst/extra-tests/`**: For tests that compare against torch.
   These live in `inst/` to avoid listing torch as a dependency.
2. **`tests/testthat/`**: For tests without a torch counterpart.

**Important**: Test names must start with `"p_<name>"` (matching the primitive object name).
The meta tests verify that every primitive has corresponding tests.

Since no torch counterpart exists for `prim_repeat_along`, we would add manual tests in:

- `tests/testthat/test-primitives-stablehlo.R`
- `tests/testthat/test-primitives-reverse.R`

Also, ensure that no linter errors are present, `devtools::check()` passes, and format the code using `make format`.

## Higher-Order Primitives

Higher-Order Primitives are primitives that parameterized by an R function or expression.
Examples include `prim_if` and `prim_while`.
These are generally much more complex to handle, so we don't cover them here in detail (for now).
The general idea, however, is that the primitive `prim_*` function needs to trace the provided function using `trace_fn()` and then forward this graph to the stablehlo lowering rule and reverse rule.
