---
title: "Adding an API Function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding an API Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will explain how to add an API function to the {anvil} package.
API functions are the main user facing API and are follow the `nv_*` naming scheme.
There are two types of API functions:

1. Thin wrappers around primitives that add convenience like type casting and scalar broadcasting (AP function `nv_add` is a wrapper around the primitive `nvl_add`).
2. Common mathematical functions that are expressed in terms of primitives (such as `nv_rnorm`).

Here, we will focus on contributing such an API function to the {anvil} package.
Specifically, API functions in the {anvil} package should:

1. Work with any backend.
2. Accept the following argument types for their dynamic inputs:
   1. `numeric` and `logical` length-1 vectors
   2. `numeric` and `logical` R arrays
   3. `AnvilArray`s
3. Always output `AnvilArray`s

## 1. Work with any backend

The first requirement means that a user of the {anvil} package should be able to use an {anvil} function with `AnvilArray`s from both the *xla* and the *quickr* backend.
When you are just calling into other anvil primitives or correctly created API functions, there is nothing to take care of.
This requirement is only relevant, when the API function is `jit()`ted (or internally calls into such a function).
For example, we might want to add an API function that adds three to a tensor and jit the function.
By setting the `backend` argument to `"auto"`, we can use it with both backends:

```{r}
library(anvil)
nv_add_three_good <- jit(\(x) x + 3, backend = "auto")
nv_add_three_good(nv_scalar(1, backend = "xla"))
nv_add_three_good(nv_scalar(1, backend = "quickr"))
```

Otherwise, the function would fail in eager mode when provided with an input from a different backend.

```{r, error = TRUE}
nv_add_three_bad <- jit(\(x) x + 3)
nv_add_three_bad(nv_scalar(1, backend = "quickr"))
```

Special care must also be taken when creating constants.
It needs to be ensured that the constant is created on the correct device (from which the backend can be determined).
If we don't specify it and do not jit the API function, the default backend is used, which means it won't work with non-default backends:

```{r, error = TRUE}
nv_add_const_bad <- function(x) {
  three <- nv_scalar(3)
  three + x
}
nv_add_const_bad(nv_scalar(1, backend = "quickr"))
```

If the API function was `jit`ted, than there would be no issue, because within a `jit()` context, the correct backend can be inferred from the jit context, so all we have to do is set `backend = "auto"`.

```{r}
nv_add_const_jit <- jit(nv_add_const_bad, backend = "auto")
nv_add_const_jit(nv_scalar(1, backend = "xla"))
```

If we don't want to JIT the API function, then we need to pass the backend of the input to the constant creation.
For that, we recommend using the `nv_<op>_like` functions that will use the attributes of the provided example as defaults for the properties of the newly created array:

```{r}
nv_add_const_eager <- function(x) {
  three <- nv_scalar_like(like = x, 3)
  x + three
}
```

## 2. Input conversion

In order to make the API function convenient to use, we auto-convert inputs.
Sometimes, this conversion needs to be done explicitly in the API function.
E.g., consider the function below that broadcasts

```{r}
nv_expand_bad <- function(x) {
  if (identical(shape, x)) {
    return(x)
  }
  nv_broadcast_to(x, shape)
}
```

* `shape_abstract`


3. Always output anvil arrays


