---
title: "Adding an API Function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding an API Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
quickr_available <- requireNamespace("quickr", quietly = TRUE)
```

Here, we will explain how to add an API function to the {anvil} package.
API functions are the main user facing API and are follow the `nv_*` naming scheme.
There are two types of API functions:

1. Thin wrappers around primitives that add convenience like type casting and scalar broadcasting (AP function `nv_add` is a wrapper around the primitive `nvl_add`).
2. Common mathematical functions that are expressed in terms of primitives (such as `nv_rnorm`).

Here, we will focus on contributing such an API function to the {anvil} package.
Specifically, API functions in the {anvil} package should:

1. Work with any backend.
2. Accept the following argument types for their dynamic inputs:
   1. `numeric` and `logical` length-1 vectors
   2. `numeric` and `logical` R arrays
   3. `AnvilArray`s
3. Always output `AnvilArray`s

## 1. Work with any backend

The first requirement means that a user of the {anvil} package should be able to use an {anvil} function with `AnvilArray`s from both the *xla* and the *quickr* backend.
When you are just calling into other anvil primitives or correctly created API functions, there is nothing to take care of.
This requirement is only relevant, when the API function is `jit()`ted (or internally calls into such a function).
For example, we might want to add an API function that adds three to a tensor and jit the function.
By setting the `backend` argument to `"auto"`, we can use it with both backends:

```{r}
library(anvil)
nv_add_three_good <- jit(\(x) x + 3, backend = "auto")
nv_add_three_good(nv_scalar(1, backend = "xla"))
```

```{r, eval = quickr_available}
nv_add_three_good(nv_scalar(1, backend = "quickr"))
```

Otherwise, the function would fail in eager mode when provided with an input from a different backend.

```{r}
nv_add_three_bad <- jit(\(x) x + 3)
```

```{r, error = TRUE, eval = quickr_available}
nv_add_three_bad(nv_scalar(1, backend = "quickr"))
```

Special care must also be taken when creating constants.
It needs to be ensured that the constant is created on the correct device and for the correct backend.
If we don't specify it and do not jit the API function, the default backend is used, which means it won't work with non-default backends:

```{r}
nv_add_const_bad <- \(x) {
  three <- nv_scalar(3)
  three + x
}
```

```{r, error = TRUE, eval = quickr_available}
nv_add_const_bad(nv_scalar(1, backend = "quickr"))
```

If the API function was `jit`ted, than there would be no issue, because within a `jit()` context, the correct backend can be inferred from the context, so all we have to do is set `backend = "auto"`.

```{r}
nv_add_const_jit <- jit(nv_add_const_bad, backend = "auto")
nv_add_const_jit(nv_scalar(1, backend = "xla"))
```

If we don't want to JIT the API function, then we need to pass the backend of the input to the constant creation.
For that, we recommend using the `nv_<op>_like` functions that will use the attributes of the provided example as defaults for `dtype`, `shape`, `backend`, `device`, and `ambiguous`:

```{r}
nv_add_const_eager <- \(x) {
  three <- nv_scalar(like = x, 3)
  x + three
}
```

The `nv_<op>_like` form is also the correct way to create constants that live on the *same device* as an arrayish input.
Calling `device()` directly on the input would work for a concrete `AnvilArray`, but under `jit()` that same input is a `GraphBox` — and `device()` on a `GraphBox` raises an error (tracing has no concrete device; jit handles placement at the input/output boundary).
Compare:

```{r, error = TRUE}
# BAD: errors under jit() because `device(x)` is not defined for a GraphBox
nv_tril_bad <- function(x) {
  zeros <- nv_fill(0, shape(x), dtype = dtype(x), device = device(x))
  nv_ifelse(x > 0, x, zeros)
}
jit(nv_tril_bad)(nv_array(matrix(c(-1, 2, -3, 4), 2, 2)))
```

```{r}
# GOOD: `nv_fill_like` reads dtype/shape/device from `x`, and under jit()
# it picks the device up from the tracing context instead of calling `device()`.
nv_tril_good <- function(x) {
  nv_ifelse(x > 0, x, nv_fill_like(x, 0))
}
jit(nv_tril_good)(nv_array(matrix(c(-1, 2, -3, 4), 2, 2)))
```

## 2. Input conversion

API functions accept R literals (length-1 vectors, R arrays), `AnvilArray`s, and (during tracing) graph boxes.
Primitives handle all of these, but having the same arrayish value arrive as three different R types makes the API function body awkward: every branch would have to re-check which case it is in.
To keep the function body simple, we standardize at the top -- after that, we can treat every argument as an `AnvilArray` (or a tracing box) without caring about the original R type.
Anvil provides two helpers for this:

* `as_anvil_array(x, device = NULL)` -- use for API functions that take a **single** array input.
  It converts an R value to an `AnvilArray` on the given device (or the default device if `NULL`), and passes `AnvilArray`s through.
* `as_anvil_arrays(...)` -- use for API functions with **multiple** array inputs.
  It infers a common device from any concrete `AnvilArray` in `...`, places R literals on that device, and errors if the concrete inputs live on different devices (or backends).

Both helpers short-circuit during tracing: inside `jit()`, device placement is already handled at the input/output boundary and when boxing constants.

### Single input

```{r}
nv_expand <- function(x, shape) {
  x <- as_anvil_array(x)
  if (identical(shape(x), shape)) {
    return(x)
  }
  nv_broadcast_to(x, shape)
}
```

### Multiple inputs

Use `as_anvil_arrays()` to normalize all arrayish inputs in one call:

```{r}
nv_add_scaled <- function(x, y, alpha) {
  args <- as_anvil_arrays(x, y, alpha)
  args[[1L]] + args[[2L]] * args[[3L]]
}
```

## 3. Always output anvil arrays

As long as you normalize the inputs with `as_anvil_array()` / `as_anvil_arrays()` and then build the result from primitives (or from values that already went through that conversion), the output is automatically an `AnvilArray`.
The one thing to watch for is short-circuit returns: if a code path returns one of the inputs unchanged (see [Return the input unchanged when possible](#return-the-input-unchanged-when-possible)), make sure the conversion step happened first so you are returning an `AnvilArray` and not the original R value.

## Inputs that require checks must be static

Any argument whose value the API function inspects (e.g. for validation, to pick a code path, or to compute a shape or `dim` index) must be a **static** argument under `jit()`.
Under tracing, arrayish inputs are replaced by `GraphBox` tracers and their concrete R value is not available -- assertions on the value would either fail or see the tracer instead of the user's data.
Static arguments, in contrast, are embedded as constants into the compiled program, so the function body sees their real R value during tracing.

As a rule of thumb, if your API function body contains `if`, `switch`, `stopifnot`, `checkmate::assert_*`, or any other code that branches on or validates a value, that argument should be static.
This typically applies to arguments like `dims`, `shape`, `dim`, flags, mode strings, or dtype specifiers -- anything that controls the *structure* of the computation rather than participating in it as data.

```r
nv_reduce_sum <- function(x, dims = NULL) {
  if (!is.null(dims)) {
    assert_integerish(dims, lower = 1L, upper = ndims(x))
  }
  # ...
}

# `dims` is inspected, so it must be static when jitted:
jit(nv_reduce_sum, static = "dims")(x, dims = 1L)
```

Arrayish inputs (the actual data being operated on) should *not* be static -- they are meant to be traced.

## Binary element-wise operations

Element-wise binary ops (`nv_add`, `nv_mul`, comparisons, ...) share a fixed preparation pipeline: promote both operands to a common dtype via `nv_promote_to_common()`, then broadcast a scalar operand up to the non-scalar shape via `nv_broadcast_scalars()`, and finally dispatch to the primitive.
The internal helper `make_do_binary()` in `R/api.R` wraps this pipeline, which is why most of the binary API functions are one-liners:

```r
make_do_binary <- function(f) {
  function(lhs, rhs) {
    args <- nv_promote_to_common(lhs, rhs)
    args <- nv_broadcast_scalars(args[[1L]], args[[2L]])
    do.call(f, args)
  }
}

nv_add <- make_do_binary(nvl_add)
```

When adding another element-wise binary primitive, reuse `make_do_binary()`.
If instead you need full NumPy-style broadcasting (not just scalar-against-tensor -- e.g. for `nv_outer()`), use `nv_broadcast_arrays()` after promotion.

## Converting auxiliary arguments to the operand dtype

Some primitives require all their inputs to share a dtype.
When your function has one "primary" operand and a few helper arguments (e.g. `min_val` and `max_val` in `nv_clamp()`, or `padding_value` in `nv_pad()`), convert the helpers to match the operand so the user can pass them as literals of a convenient type:

```r
nv_clamp <- function(min_val, operand, max_val) {
  args <- as_anvil_arrays(min_val, operand, max_val)
  op_dtype <- dtype(args[[2L]])
  nvl_clamp(
    nv_convert(args[[1L]], op_dtype),
    args[[2L]],
    nv_convert(args[[3L]], op_dtype)
  )
}
```

`nv_convert()` is itself a no-op when the dtype already matches (see next section), so the extra calls are cheap.

## Return the input unchanged when possible

Several API functions short-circuit when the requested transformation is already satisfied -- `nv_convert()` returns the input when the dtype matches, `nv_reshape()` when the shape matches, `nv_broadcast_to()` when the shape matches:

```r
nv_reshape <- function(operand, shape) {
  operand <- as_anvil_array(operand)
  if (!identical(shape(operand), shape)) {
    nvl_reshape(operand, shape)
  } else {
    operand
  }
}
```

This avoids emitting no-op graph nodes, which would otherwise clutter the computation graph and make traces harder to read.
Do this whenever the check is cheap and the no-op case is likely (common call sites, conservative defaults).

## Testing

API functions should work identically in eager mode and under `jit()`, and should preserve the device of their inputs in both modes.
The `check_eager()` helper in `tests/testthat/helper.R` checks all three properties in one call: it runs the function in eager mode with the array inputs placed on `cpu:1`, runs it `jit()`-compiled with the inputs on `cpu:0`, and asserts that

1. the eager output lives on `cpu:1`,
2. the jitted output lives on `cpu:0`, and
3. both outputs agree value-wise (within `tolerance`, default `1e-6`).

```r
test_that("nv_foo works in eager and jit mode", {
  x <- nv_array(1:6, dtype = "f32", shape = c(2, 3))
  check_eager(nv_foo, x)
})

test_that("nv_foo with a static argument", {
  x <- nv_array(1:6, dtype = "f32", shape = c(2, 3))
  check_eager(nv_foo, x, dims = 1L, static = "dims")
})
```

Forward any `static` arguments via the `static =` argument so the jit wrapper sees the same declaration you would pass to `jit()` yourself.
For non-arrayish arguments, `check_eager()` passes them through unchanged -- only `AnvilArray` inputs are placed on the test devices.

## Summary checklist

When adding a new API function, work through the following list:

- [ ] **Backend-agnostic.** The function works with both the xla and quickr backends.
  - If it creates constants, it uses `nv_*_like()` (or equivalent) so the constants pick up the backend/device from an input.
  - If it internally jit-compiles a helper, `backend = "auto"` is set on the `jit()` call.
- [ ] **Normalized inputs.** Arrayish arguments are canonicalized at the top of the function via `as_anvil_array()` (single input) or `as_anvil_arrays()` (multiple inputs).
- [ ] **Dtype promotion + broadcasting** for element-wise binary ops: go through `nv_promote_to_common()` + `nv_broadcast_scalars()`, typically via `make_do_binary()`.
- [ ] **Auxiliary args converted** to the operand dtype via `nv_convert()` when the primitive requires matching dtypes.
- [ ] **No-op shortcuts** return the input unchanged (e.g. reshape to the same shape, convert to the same dtype).
- [ ] **Static arguments** declared for every value the body inspects (shape, dims, flags, mode strings, dtype specifiers).
- [ ] **Output is always arrayish** -- short-circuit returns go through `as_anvil_array[s]()` first.
- [ ] **`_like()` variant** provided for any array-creator function.
- [ ] **Tests use `check_eager()`** to verify eager/jit parity and device preservation, with all static arguments forwarded via `static =`.
- [ ] **Exported and registered** -- `@export` tag in the roxygen block, and an entry in `_pkgdown.yml` under the appropriate section.
