---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

{anvil} supports two execution modes: *eager mode*, where each operation is compiled and dispatched on its own, and *jit mode*, where an entire function is compiled into a single executable via `jit()`.
Both modes produce identical results, but they have very different performance characteristics.
This vignette explains when to prefer each, how the compilation cache works, and how {anvil}'s asynchronous dispatch should shape the code you write around it.

If you haven't yet, read the [Get Started](anvil.html) vignette first.

```{r}
library(anvil)
```

## Eager vs. JIT

In eager mode, every `nv_*` call and every overloaded R operator compiles a tiny XLA program, submits it to the device, and returns.
This is convenient for interactive work -- you can print intermediates, step through with familiar tools, and mix {anvil} code freely with plain R.
It has two costs:

1. **Per-call overhead.** Each eager call goes through tracing, argument marshalling, and dispatch. The compilation cost is paid only once per input type (see below), but the other overheads are paid every call.
2. **No cross-operation optimization.** The XLA compiler sees only one operation at a time, so it cannot fuse consecutive operations, eliminate temporaries, or reorder work.

`jit()` compiles a whole function at once, so all of its operations end up in a single graph.
This lets XLA fuse element-wise operations, reuse buffers, and remove dead code.
For anything that chains more than a handful of operations or is called repeatedly, `jit()` is almost always the right default.

```{r}
f <- function(x) {
  y <- x * x
  z <- y + x
  z - 1
}
x <- nv_array(1:8, dtype = "f32")

# Eager: three separate dispatches
f(x)

# JIT: one fused kernel
f_jit <- jit(f)
f_jit(x)
```

Because jitted functions inline each other (see the [Composability](anvil.html#composability) section), you can freely use `nv_*` primitives inside `jit()` without paying any eager-mode overhead -- their operations are traced into the outer graph and compiled alongside the rest.

## Compilation Cache

Compilation itself isn't free -- building an XLA executable can take from milliseconds to seconds depending on the size of the graph.
{anvil} caches compiled executables keyed by the *abstract types* of the inputs: shape and dtype for `AnvilArray`s, or the value itself for arguments marked `static =`.

```{r}
g_jit <- jit(function(x, y) {
  cat("compiling ...\n")
  x + y
})
a <- nv_scalar(1, "f32")
b <- nv_scalar(2, "f32")
g_jit(a, b)  # compiles
g_jit(a, b)  # cache hit, no message
g_jit(nv_array(1:4, dtype = "f32"), nv_array(5:8, dtype = "f32"))  # recompiles
```

Two consequences follow:

- Call `jit()` once on a function and reuse the result. Calling `jit()` inside a loop creates a fresh cache on every iteration and defeats the point.
- Use `static =` sparingly. Every unique value of a static argument forces a recompilation, so reserve it for flags or small integers that only take a handful of distinct values.

See the [JIT Compilation](anvil.html#jit-compilation) section of the getting-started vignette for more on what `jit()` traces and when it recompiles.

## Asynchronous Dispatch

An `AnvilArray` is not raw data -- it wraps a device buffer that may not yet be materialized.
Both eager and jit calls return immediately with such an array; the device continues computing in the background.
While it does, R is free to prepare the next call.
This overlap is where a lot of {anvil}'s speed comes from, especially on GPUs where kernel launches and host-device transfers are expensive.

You only pay the synchronization cost when you actually need the value on the host:

- `as_array(x)` transfers the buffer back to R and returns a plain R array -- this blocks.
- `print(x)` and comparisons like `expect_equal(as_array(x), ...)` block for the same reason.
- Control flow that branches on an array value (e.g. `if (as.logical(some_flag))`) blocks to read the flag.

Metadata access -- `shape(x)`, `dtype(x)`, `device(x)` -- never blocks. It is always known from the abstract type.

This design is inherited from the underlying PJRT runtime.
The [pjrt *Get Started* article](https://r-xla.github.io/pjrt/articles/pjrt.html#asynchronous-dispatch) has a lower-level treatment, including the distinction between `PJRTBuffer` (implicit promise, passed through operations without blocking) and `PJRTArrayPromise` (explicit promise for data in transit back to R).

## Writing Efficient Loops

Async dispatch only helps if you don't accidentally serialize it.
The classic mistake is reading a value back from the device on every iteration:

```r
# Bad: blocks every step to print the loss.
for (i in seq_len(n_steps)) {
  weights <- step(X, weights, y)
  cat("loss:", as.numeric(loss(X, weights, y)), "\n")
}
```

Even though `step()` would happily run while R prepares the next iteration, the `as.numeric()` call forces R to wait for the device on every loop.
A few ways to keep the device busy:

- Log less often (every 10 or 100 steps, not every step).
- Keep the previous iteration's result around and read it *one step late*, so the current step can overlap with the previous transfer.
- Move the loop itself into `jit()` -- either as an unrolled R loop inside a jitted function, or via `nv_while()` -- so the whole loop becomes one compiled program with no host round-trips at all.

## Mutability and Donation

`AnvilArray`s are immutable and {anvil} follows value semantics (see the [Value Semantics](anvil.html#value-semantics) section of the Get Started vignette).
This raises the question of how that impacts performance.
There are two scenarios to distinguish:

1. Updating an `AnvilArray` that lives entirely within a jit-compiled function.
1. Updating an `AnvilArray` that lives in R, through a jit-compiled function.

**Within a jitted function** there is nothing to worry about.
XLA optimizes away unnecessary copies, similar to copy-on-write semantics in R: `x <- 1:10; y <- x` conceptually copies `x` on assignment, but the copy is only actually materialized when `y` or `x` is modified.
Because {anvil} compiles whole functions with shapes known at compile time, it can perform many more such optimizations.

**Across the jit boundary** -- when you assign the output of a jitted function back to an R variable -- XLA cannot see that the input arrays are no longer needed, so it has to leave them intact.
If you know that some inputs will not be needed after the call (as is the case for "update-calls" like a weight update), you can mark them as `donate` so the compiler is free to reuse their memory:

```r
update_weights_donatable <- jit(
  function(X, beta, alpha, y) {
    lr <- 0.1
    grads <- model_loss_grad(X, beta, alpha, y)
    list(
      beta = beta - lr * grads$beta,
      alpha = alpha - lr * grads$alpha
    )
  },
  donate = c("beta", "alpha")
)

weights_out <- update_weights_donatable(X, weights$beta, weights$alpha, y)
weights$beta  # errors: the buffer was donated to the call
```

This matters most in training loops over large parameters, where the alternative is allocating a fresh parameter buffer on every step.

## Further Reading

- [Get Started](anvil.html) -- introduction to {anvil}, eager/jit modes, and control flow.
- [pjrt *Get Started* article](https://r-xla.github.io/pjrt/articles/pjrt.html) -- lower-level view of the runtime, with more detail on async dispatch and writing async-friendly loops.
- [Asynchronous dispatch in JAX](https://docs.jax.dev/en/latest/async_dispatch.html) -- the design that inspired {anvil}'s execution model.
