---
title: "Working Around Static Shapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working Around Static Shapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is about a restriction that trips up most people when they first reach for `jit()`: inside a compiled function, every array's shape has to be known at trace time.
After explaining what that means in practice, we show the standard way of dealing with it -- masking -- through a handful of examples that mirror things R users reach for every day.

```{r setup}
library(anvl)
```

## The static shape restriction

When you wrap a function with `jit()`, anvl traces it once and hands the resulting graph to XLA to compile.
Every value in that graph carries a fixed shape and dtype (an `nv_aval()`, see the [JIT Deep Dive](jit.html)).
The compiler needs these up front in order to plan memory and generate code; it has no way to deal with "however many elements happen to pass the filter on the next call".

The practical consequence is that any operation whose *output shape depends on the values of its inputs* cannot be expressed as a primitive.
In R terms, this rules out the whole family of "filter by predicate" operations:

- `which(x > 0)` -- how many indices come out?
- `x[x > 0]` -- how long is the result?
- `unique(x)` -- how many distinct values are there?
- R's `table()`, `na.omit()`, `split()`, and anything else whose length you only learn by looking at the data.

None of these have a compile-time answer, so none of them can appear inside a `jit()`ed function.

A related, but separate, issue is that *different* shapes at call time land on different cache entries, so varying the shape call-to-call triggers fresh compiles.
We won't dwell on that here; the [Padding inputs to avoid recompilation](efficiency.html#padding-inputs-to-avoid-recompilation) section of the efficiency vignette covers it.
In the examples below the shape of the *input* is fixed; what varies across calls is which entries are "real".

We hope to lift the static-shape restriction in the long term -- XLA and StableHLO are gradually growing support for dynamic shapes, and the plan is for {anvl} to follow.
Until then, the patterns in this vignette are how you work around it.

## The masking pattern

The usual workaround is to keep the output shape equal to the input shape -- so that everything remains known at trace time -- and carry a logical mask that tells us which positions count.
Any downstream operation is then modified to ignore the masked-out positions.
How you do the ignoring depends on the operation:

- For a sum, masked positions contribute `0`.
- For a product, they contribute `1`.
- For a maximum, they contribute `-Inf`.
- For a conditional update, they are passed through unchanged.

The rest of this vignette is a tour of those cases.
Throughout we will use a single example vector, `x`, and compute things that in plain R we would write as `sum(x[x > 0])`, `max(x[x > 0])`, and so on:

```{r}
x <- nv_array(c(-2, 1, 3, -4, 2, -1, 5), dtype = "f32")
```

### Counting matches

The simplest case: how many entries are positive?
There is nothing to filter -- we just turn the logical vector into 0/1 and sum:

```{r}
count_positive <- jit(function(x) {
  nv_reduce_sum(nv_convert(x > 0, "i32"), dims = 1L)
})

count_positive(x)
```

This is the building block for everything that follows.
Whenever you are tempted to reach for `length(x[cond])`, this is the replacement.

### Masked sum and mean

For a sum, the neutral value is `0`, so zeroing the masked-out entries and summing the whole vector gives the right answer:

```{r}
sum_positive <- jit(function(x) {
  nv_reduce_sum(nv_ifelse(x > 0, x, nv_fill_like(x, 0)), dims = 1L)
})

sum_positive(x)
```

We use `nv_fill_like()` to make the `0` match `x`'s dtype, shape, and device -- see the [Creating constants on the right device](correct_functions.html#creating-constants-on-the-right-device) section of the correct-functions vignette.
Multiplying by the mask works too (`x * nv_convert(x > 0, dtype(x))`), with essentially the same result.

The mean is where the output-length problem becomes visible.
`mean(x[x > 0])` needs both the sum *and* the count, and those counts vary with the data -- but because each is a scalar we can compute them separately and divide:

```{r}
mean_positive <- jit(function(x) {
  mask <- x > 0
  total <- nv_reduce_sum(nv_ifelse(mask, x, nv_fill_like(x, 0)), dims = 1L)
  n <- nv_reduce_sum(nv_convert(mask, "f32"), dims = 1L)
  total / n
})

mean_positive(x)
```

Note that the divisor is the number of matching entries, not `length(x)`.
This is the same trick used for padded inputs in the efficiency vignette; the difference is only that here the "invalid" positions are picked by a predicate on the data, rather than by a length argument.

### Masked max and min

A maximum over a subset needs a different neutral value: whatever is subtracted from the input won't change, so we want something *smaller than anything real*.
For a floating-point reduction that is `-Inf`:

```{r}
max_positive <- jit(function(x) {
  nv_reduce_max(nv_ifelse(x > 0, x, nv_fill_like(x, -Inf)), dims = 1L)
})

max_positive(x)
```

Symmetrically, `nv_reduce_min()` wants `+Inf` as the fill.
For integer dtypes, `-Inf` / `+Inf` are not representable; use the dtype's minimum and maximum (`.Machine$integer.max`, `-.Machine$integer.max`, and analogues for the anvl dtype you are working with).

A practical caveat: if it is possible that *no* entries pass the predicate, `nv_reduce_max()` will return `-Inf` rather than erroring.
In plain R, `max(numeric(0))` also returns `-Inf` (with a warning), so the behaviours line up -- but if you want an actual error you have to check the count yourself.

### Selective update

Not every use of `x[cond]` is about filtering; a lot of R code uses it to update only the matching entries, for instance clipping negatives to zero:

```r
x[x < 0] <- 0
```

Here the output has the same shape as the input, so there is no filtering problem at all.
The direct translation is `nv_ifelse()`:

```{r}
clip_negative <- jit(function(x) {
  nv_ifelse(x < 0, nv_fill_like(x, 0), x)
})

clip_negative(x)
```

This is worth flagging because, written with `[<-`, it *looks* like it should be a problem but is not.
Whenever the update does not change the array's shape, you can do it inside `jit()`.

### Masked variance

The last example is where just "fill with zero and go" stops being enough.
Variance and standard deviation involve both the mean and the spread, and the count enters twice -- once in the mean subtraction, once in the denominator.
A naive "fill with 0, then call `nv_var()`" would pull all the masked-out zeros into both the mean and the deviation, silently giving you the wrong answer.
The fix is to compute the pieces by hand, using the true count in both places:

```{r}
var_positive <- jit(function(x) {
  mask <- x > 0
  mask_f <- nv_convert(mask, "f32")
  n <- nv_reduce_sum(mask_f, dims = 1L)
  mean_x <- nv_reduce_sum(nv_ifelse(mask, x, nv_fill_like(x, 0)), dims = 1L) / n
  deviations <- nv_ifelse(mask, x - mean_x, nv_fill_like(x, 0))
  nv_reduce_sum(deviations * deviations, dims = 1L) / (n - 1)
})

var_positive(x)
var(as_array(x)[as_array(x) > 0])
```

The pattern here is general: any statistic whose formula involves a count needs the *true* count rather than the padded length, and any intermediate quantity that is multiplied or summed needs the masked-out positions to contribute the appropriate neutral value.
Once you have internalised that, most "summary over a filtered subset" computations become mechanical.

## When masking is not enough

Masking works when the final answer has a shape that is fixed at trace time -- a scalar, a length-`k` summary, one value per row of a matrix.
It does not help when the thing you actually want to produce is itself of data-dependent length: the indices returned by `which(x > 0)`, the distinct values returned by `unique(x)`, the truncated vector returned by `na.omit(x)`.
There is no representation for those inside the compiled graph, and no masking trick that gets around it.

The practical options are:

- **Keep the data-dependent step outside `jit()`.** Compute the indices or the distinct values in plain R (or in eager `nv_*`), and pass the result in as a fresh input to the next jitted call. The downside is a recompile every time the resulting shape changes.
- **Reformulate the algorithm so the output shape is static.** Instead of returning the elements that match a predicate, return a summary of them (as in the examples above), or a fixed-length representation that encodes the match (e.g. `x * (x > 0)` keeps a length-`n` vector in which the unwanted positions are already zeroed).
- **Pad to a known upper bound and carry a mask.** If you know the answer has at most `k` entries for some fixed `k`, you can often produce a length-`k` result plus a length-`k` mask. This is how JAX handles the equivalent operations (`jax.numpy.unique(..., size = k)`), and the same idea works here.

Most of the time, reformulating is the right move: the filtering was never the end goal, only a step towards some statistic, and computing that statistic directly with a mask is both simpler and faster than pretending filtering is available.
