---
title: "Subsetting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, you will learn how to subset tensors in {anvil} and how to update subsets.
Because tensor shapes in {anvil} programs are static (for now), only certain subsetting operations are supported and they come with some surprises.

First, it is only possible to subset via `indices` and not, for example, using boolean masks.
This is because the output shape of a tensor subsetted by a boolean mask is unknown as it depends on how many elements in the boolean mask are `TRUE`.

```{r, error = TRUE}
library(anvil)
jit_eval({
  x <- nv_tensor(1:3)
  x[nv_tensor(c(TRUE, TRUE, FALSE))]
})
```

Second, no error can be thrown when subsetting with invalid indices, as the XLA backend we use for compilation does not support this.
Instead, out-of-bound indices are clamped to the valid range:

```{r}
jit_eval({
  x <- nv_tensor(1:3)
  x[nv_tensor(3:5, dtype = "i32")]
})
```

## Index Types

With these restrictions in mind, we will now cover the different ways to subset a tensor.
A complete subset is defined by specifying an individual index for each dimension of a tensor.
Such an index can be:

1. A **known single element** (R integer literal), which drops the dimension.
2. A **known range** (e.g., `2:5`), which preserves the dimension.
3. A **known list** (e.g., `list(1, 3, 6)`), which selects non-contiguous elements and preserves the dimension.
4. An **unknown single element** (0-dimensional tensor), which drops the dimension.
5. An **unknown list** (1-dimensional tensor), which preserves the dimension.

These can be combined freely across different dimensions.

## Dropping vs. Preserving Dimensions

Let's start with a 1D tensor:

```{r}
x <- nv_tensor(1:10, dtype = "i32")
```

If we want to get a single element, we have to decide whether to drop the dimension or not.
A scalar R index drops the dimension, while `list()` with a single element preserves it:

```{r}
# Drops the dimension (result is a scalar)
jit_eval(x[1])

# Preserves the dimension (result has shape (1))
jit_eval(x[list(1)])
```

The same distinction applies to tensor indices: a 0D scalar tensor drops the dimension, while a 1D tensor of shape `(1)` preserves it:

```{r}
# Drops the dimension
jit_eval(x[nv_scalar(1L)])

# Preserves the dimension
jit_eval(x[nv_tensor(1L, dtype = "i32", shape = 1L)])
```

## Ranges

Ranges always preserve dimensions and select contiguous slices:

```{r}
jit_eval(x[2:5])
```

Ranges must be static (known at compile time).

## Selecting Multiple Non-Contiguous Elements

Use `list()` to select multiple non-contiguous elements:

```{r}
jit_eval(x[list(1, 5, 10)])
```

Alternatively, pass a 1D tensor:

```{r}
jit_eval(x[nv_tensor(c(1L, 5L, 10L), dtype = "i32")])
```

Note that R vectors of length > 1 are not supported as subset indices, because they can't be distinguished from scalar literals in all cases.
Use `list()` for known indices and 1D tensors for dynamic indices.

## Multi-Dimensional Subsetting

For multi-dimensional tensors, specify one index per dimension separated by commas.
Trailing dimensions that are left unspecified select all elements:

```{r}
m <- nv_tensor(matrix(1:12, nrow = 3, ncol = 4), dtype = "i32")

# Select row 2 (drops dimension 1)
jit_eval(m[2, ])

# Select column 3 (drops dimension 2)
jit_eval(m[, 3])

# Select a submatrix (2:3 rows, 1:2 columns)
jit_eval(m[2:3, 1:2])

# Trailing `,` is optional: m[2:3] is the same as m[2:3, ]
jit_eval(m[2:3])
```

You can also select non-contiguous elements along multiple dimensions simultaneously.
This creates a cartesian product of the indices:

```{r}
# Select rows 1 and 3, columns 2 and 4 → result has shape (2, 2)
jit_eval(m[list(1, 3), list(2, 4)])
```

## Subset Assignment

Use `nv_subset_assign()` or `[<-` to update elements of a tensor.
This has copy-on-write semantics: a new tensor is returned with the specified positions updated.

```{r}
jit_eval({
  x <- nv_tensor(1:10, dtype = "f32")
  x[3:5] <- nv_tensor(c(30, 40, 50), dtype = "f32")
  x
})
```

Scalar values are broadcast to fill the subset:

```{r}
jit_eval({
  x <- nv_tensor(1:10, dtype = "f32")
  x[3:5] <- nv_scalar(0, dtype = "f32")
  x
})
```

Multi-dimensional assignment works the same way:

```{r}
jit_eval({
  m <- nv_tensor(matrix(0, 3, 4), dtype = "f32")
  m[1:2, 2:3] <- nv_tensor(matrix(c(1, 2, 3, 4), 2, 2), dtype = "f32")
  m
})
```

## Differentiability

Both `nv_subset()` and `nv_subset_assign()` are differentiable.
Gradients flow through the subset operations correctly:

```{r}
grad_fn <- jit(gradient(function(x) {
  nv_reduce_sum(x[2:4])
}))
grad_fn(nv_tensor(1:5, dtype = "f32"))
```

The gradient is 1 at positions 2-4 (the elements that contribute to the sum) and 0 elsewhere.

For subset assignment, gradients flow to both the original tensor and the assigned value:

```{r}
grad_fn <- jit(gradient(function(x, v) {
  y <- x
  y[2:3] <- v
  nv_reduce_sum(y)
}))

grad_fn(
  nv_tensor(1:5, dtype = "f32"),
  nv_tensor(c(10, 20), dtype = "f32")
)
```

Positions that were overwritten in `x` receive gradient 0, while the assigned values in `v` receive gradient 1.
