---
title: "anvl: Fast Array Computing and Code Transformations for R"
date: "2025-12-09"
abstract: > # max 250 words
  The \CRANpkg{anvl} package is an array computing framework for R with support for just in time (jit)
  compilation, execution on accelerators (such as GPUs) and reverse-mode automatic differentiation.
  The goal of the package is to simplify the development of new statistical algorithms
  in R and to allow them to build scalable numerical algorithms in pure R without having to
  drop down to compiled languages like C++.
  Also, it can serve as a basis for re-implementing existing functionality of compute-intensive
  R packages (e.g. for Bayesian algorithms) to better leverage accelerators such as GPUs.
  In the R ecosystem, \CRANpkg{anvl} is most closely related to \CRANpkg{torch}, but whereas
  \CRANpkg{torch} is primarily designed around eager execution, \CRANpkg{anvl} is designed around
  just-in-time compilation, which can yield considerably better performance, e.g. for workloads
  dominated by many small operations rather than a few large kernels.

  \CRANpkg{anvl} achieves this via *code transformations*: functions that take an R function as
  input and return a new R function --- reverse-mode automatic differentiation, for example, maps a
  function to its gradient.
  Internally, such transformations are implemented by tracing R functions into an intermediate
  representation; the primary execution backend is built on OpenXLA, in particular its StableHLO
  compiler infrastructure.
  By separating the *what* of an algorithm from the *how* of its execution, anvl allows for
  connecting alternative backends, one example being the experimental \CRANpkg{quickr} backend, which
  transpiles to Fortran.
  Apart from these backends, \CRANpkg{anvl} is written in R, which together with its modular
  design makes it easy to inspect, extend, and contribute to, e.g.\ for adding new primitives or
  code transformations.
draft: true
author:
  - name: Sebastian Fischer
    affiliation:
    - LMU Munich
    - Munich Center for Machine Learning
    address:
    - Department of Statistics
    - Ludwig-Maximilians-Universität Munich
    - Ludwigstr. 33, 80539 Munich, Germany
    orcid: 0000-0002-9609-3197
    email:  sebastian.fischer@stat.uni-muenchen.de
  - name: Daniel Falbel
    affiliation:
    - Posit Software, PBC
    orcid: 0000-0002-0912-0225
  - name: Thomas Kalinowski
    affiliation:
    - Posit Software, PBC
  - name: Niko German
    affiliation:
    - LMU Munich
    orcid: 0009-0001-7394-8367
  - name: Martin Binder
    affiliation:
    - LMU Munich
    - Munich Center for Machine Learning
    orcid: 0009-0008-2578-2869
  - name: Bernd Bischl
    affiliation:
    - LMU Munich
    - Munich Center for Machine Learning
    orcid: 0000-0001-6002-6980
type: package
output:
  rjtools::rjournal_pdf_article:
    toc: no
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: references.bib
filters:
  - fix-macros.lua

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```



\input{notation.tex}

Final iteration

* [ ] check that references are correct
* [ ] remove lua hook for final submission
* [ ] Authors:
  * [ ] T Kalinowski (ask)
  * [ ] Bernd
  * [ ] Martin
  * [ ] Marc

TODOs:

* [x] mention R's vectorize() --> similar functional spirit as {anvl}
* [x] stress that {quickr} is experimental and not really useable yet.
* [x] mention interoperability: pjrt can run other {stablehlo} programs
* [ ] good benchmark comparing HMC from rstan with pure {anvl} implementation (also show GPU use).
* [x] mention which bayesian frameworks are built on top of JAX in Python
* [x] mention that we test against torch for autodiff
* [x] future work:
  * [x] jacobians
  * [x] hessians
  * [x] proper dynamism support with quickr backend
* [x] Stress that {anvl} is also about offering a more consitent API for array manipulating (to distinguish with {quickr})
* [x] Discuss eager and jit mode
* [x] Talk about async
* [ ] CRAN submision: Download of prebuilt binaries
* [ ] Really think about possible license issues
* [ ] include benchmark for tracing / compiler overhead and discuss:
  * [ ] despite compiler overhead --> better scaling to expensive problems / large datases
* [ ] Create table that really does proper comparison with related work w.r.t. different features of the package (include in appendix maybe).
* [ ] compare with stan in more depth (AD engine; transpilation vs. proper XLA compiler)
* [ ] discuss compile-time of quickr (2s)
* [ ] use deep-research again for related work
  * [x] separating how from what
* [x] anvl really low level infrastructure as a basis for others to develop new algorithms.
* [ ] static vs tracing:
    * [ ] tracing requires control of the R function (now rowSums etc.)
    * [ ] tracing allows to "borow" more from the host interpreter (static arguments)
* [x] related work:
  * [x] OpenCL package by Simon Urbanek
  * [x] discuss problems: heterogenous compute
* [x] stablehlo / pjrt in itself: interoperability with models from other programming languages.




## Introduction

<!-- High-level Motivation -->
All statistical and machine learning models fundamentally rely on some form of vector or array operations.
Even though R as an interpreted language has performance limitations, relying on optimized matrix operations can often be fast enough for many use cases.
However, there are cases where the performance is not sufficient and one needs to resort to compiled languages such as C++.
Furthermore, one might additionally want to have access to GPU acceleration or automatic differentiation (AD) for gradient-based optimization.

<!-- Contribution -->
To address this, we introduce \CRANpkg{anvl}, which is a composable code transformation framework for the R language.
R is no stranger to such transformations: base R already ships `Vectorize()`, which lifts a scalar function to a vectorized one and is similar in spirit to batch-map transformations such as `vmap` in JAX and PyTorch.
Its primary goal is to offer a set of primitive operations that can be jit-compiled, differentiated, and executed on different hardware backends.
A second, equally important goal is to provide an API that is well-suited to working with higher-dimensional arrays, so that users can focus on expressing *what* an algorithm computes rather than on the bookkeeping of *how* it is implemented in terms of R's array semantics.
It therefore allows users to implement numerical algorithms efficiently in pure R.
Two different execution backends are supported: an OpenXLA ("Open Accelerated Linear Algebra") backend and an experimental Fortran backend.
The OpenXLA project is a compiler ecosystem used across different array computing frameworks.
Part of this project are the StableHLO language, which is a strongly typed language for representing array programs, and the PJRT (Pluggable Jit RunTime) library, which provides a common runtime for executing compiled programs.
Components of \CRANpkg{anvl} are the \CRANpkg{stablehlo} and \CRANpkg{pjrt} packages.
The former allows users to easily create StableHLO programs in R, while the latter interfaces the PJRT C library and handles compilation, data allocation, and execution.
Besides the OpenXLA backend, there is also an alternative, experimental backend that transpiles R to Fortran code via the \CRANpkg{quickr} package.
Because the \CRANpkg{anvl} package is primarily written in R, it is easy to extend and contribute to.

<!-- Related work -->

<!-- Inspirations for design -->

The closest related work is the JAX python library, from which various design elements are inspired or adopted [@ref-jax].
Primarily, we are using a similar functional design and handle jit compilation via OpenXLA.
Also other design elements such as the handling of nested data structures or the type promotion rules are adopted from JAX.
We also want to highlight the Autodidax tutorial, which describes the core ideas underpinning JAX (todo: cite), as well as microjax, which is a minimalistic Python library demonstrating JAX's code transformation engine (todo: cite).
For the design of the \CRANpkg{stablehlo} package and the tracing implementation in \CRANpkg{anvl}, we have followed the design from \CRANpkg{mlr3torch}, which offers a similar graph-building interface for creating neural networks.
The API functions for the array operations are partly adopted from NumPy, which is a popular library for numerical computing in Python [@ref-harris2020array].
For the design of the \CRANpkg{pjrt} package, we have partly borrowed from the gopjrt library [@ref-gopjrt] which offers a go interface to PJRT.

<!-- OpenXLA, stablehlo, pjrt-->

The OpenXLA project [@ref-openxla] is an open-source compiler ecosystem for machine learning, originally developed at Google as part of TensorFlow and later released as a standalone project.
One of its goals is to decouple numerical computing frameworks (the frontends) from hardware-specific implementations (the backends) by defining two standardized interfaces.
One of these is the StableHLO language [@ref-stablehlo], a portable intermediate representation for array programs.
StableHLO is built on MLIR (Multi-Level Intermediate Representation), which is itself part of the LLVM compiler infrastructure.
MLIR provides a framework for defining and transforming intermediate representations at multiple levels of abstraction that are transformed into each other via passes that e.g. optimize the program or lower the abstraction level.
The second interface is PJRT (Pluggable JIT RunTime), which defines a common API for the essential operations of a JIT-compiled array computing framework such as buffer management, compilation, and execution.
PJRT follows a plugin architecture, where each vendor can provide a shared library that implements the PJRT interface for their specific hardware.
This means that any framework that emits StableHLO and uses PJRT for execution can automatically target any hardware for which a PJRT plugin exists, including CPUs, NVIDIA GPUs, and Google TPUs.
Currently, the main compiler consuming StableHLO programs is the XLA compiler, but alternatives such as the IREE compiler [@ref-iree] also exist.

<!-- Difference between anvl and torch: More technical motivation -->

The main motivation behind developing \CRANpkg{anvl} is to go beyond the limitations of the \CRANpkg{torch} package [@ref-rtorch], which is a deep learning framework for the R language.
It is based on LibTorch, which is the C++ library underpinning PyTorch [@ref-pytorch].
The \CRANpkg{torch} package offers a collection of highly optimized array operations, as well as various modules for building and training neural networks.
As such, it also has support for backward-mode automatic differentiation and can be run on different hardware backends, including CPU and GPU.
Thus, its general capabilities are similar to those of \CRANpkg{anvl}, although it is more mature and contains more features.
However, \CRANpkg{torch} is primarily designed around so-called "eager" execution.
For example, the forward function of a neural network is simply an R function that repeatedly calls into LibTorch kernels; i.e., there is no compilation step.
\CRANpkg{anvl} supports both modes: by default, primitive operations execute eagerly just like in \CRANpkg{torch}, and JIT compilation is opted into by wrapping a function in `jit()`.
While eager execution has advantages regarding usability and expressiveness, it comes with performance implications.
For large neural networks that are in the primary focus of modern deep learning toolboxes, these performance implications might be less problematic, as the runtime is dominated by individually expensive operations such as the attention mechanism.
However, when models are smaller---as is not unusual for statistical models---JIT compilation can be especially attractive.
A second, related advantage of the JIT-compilation approach is that it gives end-users substantially more flexibility to customize their algorithms.
Consider Gaussian process regression as an example: in eager-execution frameworks, users tend to be restricted to the set of kernel functions that the underlying library ships with, since expressing a custom kernel in pure R would incur per-operation interpreter overhead on every evaluation.
With JIT compilation, in contrast, custom kernels written in pure R are compiled to efficient code and run at the same speed as the built-in ones.
While torch offers a way to JIT-compile code via tracing code and converting it to TorchScript, this feature is in maintenance mode.
Furthermore, it is primarily designed for deployment and has, for example, no support for automatic differentiation.

<!-- DL frameworks -->

Besides \CRANpkg{torch}, other R packages exist that offer GPU acceleration and automatic differentiation.
These are \CRANpkg{tensorflow} and \CRANpkg{keras3}, two deep learning frameworks that interface the TensorFlow Python runtime via \CRANpkg{reticulate}.
It is also possible to use Python's JAX library via \CRANpkg{reticulate} [@ref-jax].
All of these packages also rely on OpenXLA for execution.
That XLA compilation is a valuable tool for implementing statistical algorithms is further evidenced by the popularity of JAX-based libraries such as NumPyro [@ref-numpyro] and BlackJAX [@ref-blackjax], which drive much of the Bayesian modeling and inference ecosystem in Python.
While they provide access to more mature frameworks than \CRANpkg{anvl}, they require managing the Python dependency, as well as type conversions between R and Python.
Furthermore, with \CRANpkg{anvl} our goal is to develop an array computing library that is and will be developed primarily with the needs of the R community in mind.

Beyond those popular deep learning libraries, other R packages exist that offer hardware acceleration, automatic differentiation, jit compilation, or a combination of these.
We here primarily focus on those packages that are either available from standard repositories such as CRAN or Bioconductor, or which seem to be ready to be used.

<!-- JIT compilation -->
A recent package that enables JIT compilation in R is the \CRANpkg{quickr} package, which transpiles R code into Fortran code and then compiles it [@ref-quickr].
There are various differences between \CRANpkg{anvl} and \CRANpkg{quickr}.
First, \CRANpkg{quickr} has no support for automatic differentiation and currently only runs on CPU.
Besides, \CRANpkg{quickr} does not establish a new API, but instead aims to mirror the functionality of base R as closely as possible and act as a drop-in replacement for a subset of R code.
\CRANpkg{anvl}, in contrast, deliberately introduces a dedicated array-computing API that is designed to make working with higher-dimensional arrays more ergonomic and to separate *what* an algorithm computes from *how* it would otherwise have to be expressed in terms of R's array semantics.
This works because it is implemented via a static analysis approach, so the package does not need control over the functions that can be compiled, as is the case for the tracing approach followed in \CRANpkg{anvl}.
This also means that \CRANpkg{quickr} does not allow mixing "non-compilable" R code with compilable code, as is possible in \CRANpkg{anvl}, where the "non-compilable" code is staged out during tracing to the intermediate representation.
The tracing approach is more flexible, but also requires the user to be more careful about the code that they are writing.
There is also the ast2ast package, which translates a subset of R code into C++ [@ref-ast2ast].
It is similar in spirit to \CRANpkg{quickr}, but also supports forward and backward-mode automatic differentiation.
Then, there are \CRANpkg{odin} and \CRANpkg{rxode2}, both of which focus on differential equations and compile a small subset of R code via translation to C, similar to \CRANpkg{quickr}.
However, they do not support GPU acceleration and their set of supported operations is more limited than \CRANpkg{anvl}'s.
There is another category of packages that offer special DSLs for statistical models which are compiled.
For example, \CRANpkg{brms} [@ref-brms] allows users to define Bayesian models via the familiar formula interface and translates them into Stan models which are compiled [@ref-stan].
Stan itself is a probabilistic programming language and also has GPU support via OpenCL [@ref-opencl].
It is also possible to directly use Stan in R, e.g. via the \CRANpkg{rstan} package, but this requires writing the program itself in Stan.
There also exists \CRANpkg{nimble}, which is primarily intended for fitting hierarchical statistical models and works by generating C++ code and compiling it.
It also supports compiling custom R functions by generating C++ code, as well as forward and backward-mode automatic differentiation via CppAD [@ref-cppad].
However, there is no support for GPU acceleration.
Then, there is also the \CRANpkg{compiler} package of the R language itself [@ref-r].
It is used by the R interpreter to compile arbitrary R code into byte-code that is executed by the R virtual machine.
However, because it has to handle arbitrary R code, the possible optimizations are limited due to R's design [@ref-evaluating-r].
This generally highlights the trade-off between expressiveness and performance.
In \CRANpkg{anvl}, we are focusing on a rather narrow set of numerical operations, which thus allows for aggressive optimizations and highly efficient execution.

<!-- More Hardware acceleration -->

While various R packages offer GPU acceleration for specific operations (e.g., \CRANpkg{xgboost}), generic GPU-accelerated array computing frameworks beyond the deep learning libraries from earlier are rare.
OpenCL is an open standard for heterogeneous parallel computing across CPUs, GPUs, and other accelerators, in which kernels are written in a C-like language and JIT-compiled by the device's vendor runtime.
One noteworthy mention is the \CRANpkg{OpenCL} package [@ref-r-opencl], which exposes a low-level interface to OpenCL and requires users to hand-write and dispatch kernels themselves, offering no JIT compilation of R code or automatic differentiation.
It also leaves the OpenCL runtime setup to the user, who must install a vendor ICD separately, whereas \CRANpkg{anvl} ships prebuilt PJRT plugins that bundle the required runtime libraries.

<!-- Automatic Differentiation -->

Besides the packages already listed above, other R libraries exist with automatic differentiation capabilities.
For example, forward mode automatic differentiation via dual numbers is supported by the packages \CRANpkg{nabla}, \CRANpkg{salad}, \CRANpkg{dual}, and \CRANpkg{madness} [@ref-nabla, @ref-salad, @ref-dual, @ref-madness].
None of these packages have JIT compilation capabilities or support GPU execution.

There is also the \CRANpkg{RTMB} package which allows for forward and backward-mode automatic differentiation and interfaces the TMB C++ library [@ref-tmb].

Despite many packages offering some subset of \CRANpkg{anvl}'s features, none of the native R packages simultaneously offer GPU acceleration, JIT compilation, and backward-mode automatic differentiation.

The structure of the paper is as follows:
First, we will briefly review some important background information, including the OpenXLA compiler, the \CRANpkg{quickr}, and some basics on automatic differentiation.
In the subsequent part, we will describe the main design of the \CRANpkg{anvl} package.
Then, we will demonstrate the package's abilities, using logistic regression as a simple example.
Afterwards, we will show how it is possible to extend the package.
Then, we will include some small CPU runtime benchmarks to illustrate the performance of the package.
Finally, we will conclude by discussing current limitations and directions for future work.

## Background

### The OpenXLA Compiler Ecosystem

The OpenXLA compiler ecosystem consists of various individual components, but we will only cover the StableHLO intermediate language and the PJRT library which are used directly in \CRANpkg{anvl}.
One of the main goals of the OpenXLA project is to establish standardized interfaces between numerical computing frameworks such as JAX or anvl and their underlying execution engines.
This is achieved via the following two components:

1. The "Pluggable Jit RunTime" (PJRT) which defines a common interface for execution engines.
2. The StableHLO language, which is the intermediate representation produced by the frontend.

If a numerical computing framework produces StableHLO programs and handles the execution via PJRT, it can be run on any hardware for which a PJRT plugin exists.
This turns an $n \times m$ problem ($n$ frontends need to implement their execution for $m$ different hardware backends) into a $n + m$ problem.

#### StableHLO

An example for the StableHLO intermediate representation is given below.
The main program takes in two `f32` arrays of shape `2x2`, multiplies them, and returns the result.

```mlir
func @main(%arg0: tensor<2x2xf32>, %arg1: tensor<2x2xf32>) -> tensor<2x2xf32> {
  %0 = "stablehlo.multiply"(%arg0, %arg1) : (tensor<2x2xf32>, tensor<2x2xf32>) -> tensor<2x2xf32>
  return %0 : tensor<2x2xf32>
}
```

It is expressed in MLIR (Multi-Level Intermediate Representation).
MLIR is part of the LLVM compiler infrastructure and provides a framework to define intermediate representations at multiple levels, also called dialects, that are transformed into each other via so-called passes.
Such passes can e.g. perform optimizations or lower the abstraction level.
It is in single static assignment (SSA) form, meaning that each variable is assigned to exactly once, and is fully functional.
For a description of the language, see the [StableHLO specification](https://openxla.org/stablehlo/spec), which also contains a description of the around 100 supported primitive operations.
In order to use the StableHLO language in \CRANpkg{anvl}, it needs to be possible to translate R code into StableHLO programs.
For this purpose, we have created the \CRANpkg{stablehlo} package.
The package essentially consists of four parts:

1. An S3 class representation of the StableHLO language, which closely follows the specification.
2. A `repr()` generic that converts the class representation into a string representation that can be compiled by the XLA compiler.
3. A set of type inference functions for the primitive operations.
4. A builder API for easily building up StableHLO programs.

The package is implemented fully in R, instead of interfacing the C++ implementation, which makes it easy to install for end users.
However, the \CRANpkg{stablehlo} package is only a partial implementation of the language and does not support all features.
For example, there is no support for complex numbers, a few primitive operations are missing, and there is currently no support for shape dynamism (unknown dimensions).

The example below illustrates how to use the builder API to re-create the StableHLO program from above.

```{r}
library(stablehlo)
func <- local_func("main")
func
x <- hlo_input("x", "f32", shape = c(2, 2), func = func)
x
y <- hlo_input("y", "f32", shape = c(2, 2), func = func)
z <- hlo_multiply(x, y)
f <- hlo_return(z)
f
hlo_string <- repr(f)
```

What happens in the background is that first, a new `Func` object is created with the name "main", which has reference semantics.
Then, `hlo_input()` creates two `FuncInput` objects with the names "x" and "y" and specifies their type.
These are added to the `$inputs` field of the built-up `Func` object.
The return values are `FuncValue` objects, which represent variables in the underlying StableHLO program that is being built up.
Objects of type `FuncValue` represent a specific value in a StableHLO program and are the core data structure underpinning the builder API.
Note that the R variable name `x` that is bound to the `FuncValue` does not need to coincide with the identifier "x" of the value in the StableHLO program.
Further note that the `func` object here is passed explicitly, but this would not be necessary as the default is to use the function that was last created.
When `x` and `y` are multiplied via `hlo_multiply()`, the corresponding type inference function `stablehlo::infer_types_multiply()` is called, which checks that the operation is valid for the given input types, computes the output type of the operation, registers the operation in the body of the underlying StableHLO program, and then
returns a `FuncValue` object for the result.
Then, `hlo_return()` sets the return values in the underlying `Func` object and returns the `Func` object itself, marking the end of the program creation.

Finally, the `repr()` generic is called on the `Func` object to get the string representation of the program.
This is implemented by recursively calling the `repr()` generic on its sub-components.
Note that the printer for `Func` and `FuncValue` objects uses the `repr()` generic internally to print the sub-components.

While some details were omitted in the above example, this shows the core ability of the \CRANpkg{stablehlo} package to easily create StableHLO programs in R.
For some more details, see the StableHLO specification or [this vignette](https://r-xla.github.io/stablehlo/articles/stablehlo.html).

#### PJRT

The \CRANpkg{pjrt} package is the interface to the PJRT C++ library.
The PJRT library itself is header-only and only defines the interface for the underlying execution engine.
A specific implementation of the interface (e.g. for CPU or NVIDIA GPU) is accessed by downloading prebuilt shared libraries and loading them dynamically.
Core components of the PJRT interface are buffer allocation, program compilation, and program execution, but many other API endpoints exist.

A practical consequence of this plugin architecture is that GPU support becomes considerably easier to install than with traditional CUDA-based R packages.
The CUDA PJRT plugin is self-contained: it ships with the required CUDA runtime libraries, so the user only needs a compatible NVIDIA driver and does not have to install a matching CUDA toolkit on the host system.

Because \CRANpkg{pjrt} only requires a StableHLO program as input, it is not tied to \CRANpkg{anvl} as a frontend.
Models trained in other frameworks --- for example PyTorch via `torch.export`, JAX via `jax.export`, or TensorFlow via the StableHLO converter --- can be exported to StableHLO and then loaded and executed from R through \CRANpkg{pjrt}, opening up a path to deploy and run pre-trained models in R workflows without depending on the original Python toolchain.
The reverse direction is equally interesting: StableHLO programs produced by \CRANpkg{anvl} can in principle be exported and run by any other StableHLO-consuming framework or runtime.
Realizing both directions in practice will require extending the \CRANpkg{stablehlo} package to support programs with dynamic dimensions and the corresponding shape-refinement passes, since exported programs typically use shape polymorphism that has to be specialized down to concrete shapes before XLA compilation; this is not yet supported.

The same dependency scale is also the reason why \CRANpkg{pjrt} ships prebuilt plugins rather than building them from source at install time.
A PJRT plugin transitively depends on LLVM, MLIR, the StableHLO and XLA dialects, parts of the XLA runtime, and further libraries such as protobuf, Abseil, and Eigen, adding up to a rather complicated Bazel-driven build that produces gigabytes of artifacts.
On the GitHub Actions runners we use for CI, a full plugin build takes between one and four hours, so expecting end users to run this as part of a source install would be unreasonable.
In addition, separate plugins are built for CPU and CUDA today (with Metal likely to follow), each targeting several operating systems, so building from source on the user's machine would mean re-running this process for every platform-backend combination.

Just like for the \CRANpkg{stablehlo} package, we only show the core functionality of \CRANpkg{pjrt} here and refer to the corresponding [vignette](https://r-xla.github.io/pjrt/articles/pjrt.html) for more details.

As a first step, we will compile the StableHLO program from above.
For this, the R `character(1)` first needs to be converted into a `PJRTProgram` object.

```{r}
library(pjrt)
program <- pjrt_program(src = hlo_string, format = "mlir")
```

Then, this `PJRTProgram` can be compiled into a `PJRTLoadedExecutable` object, targeting a specific device.
A `PJRTDevice` is obtained from a `PJRTClient`, which in turn wraps a `PJRTPlugin` --- the dynamic library implementing the PJRT interface for a given hardware backend.
Each platform has a single cached client, constructed on demand from a string identifier such as `"cpu"` or `"cuda"`.

```{r}
device <- pjrt_device("cpu")
executable <- pjrt_compile(program, device = device)
executable
```

In order to execute this program, `PJRTBuffer` objects need to be created that are multi-dimensional arrays.
Below, we create two simple `f32` arrays of shape `2x2` and specify the device to be the CPU, which will use the buffer allocation implementation of the PJRT CPU plugin.
For CPU, the distinction between device and client is not so important as there is usually only one CPU, so there is a one-to-one correspondence.
When working with GPUs, there might be multiple GPUs available and the device represents one of them (by default the first one), and it matters on which GPU the buffer will be allocated.
However, multi-GPU setups are not properly supported in \CRANpkg{pjrt} yet, but this could be added in the future.

```{r}
x <- pjrt_buffer(1:4, dtype = "f32", device = "cpu")
x
y <- pjrt_buffer(5:8, dtype = "f32", device = "cpu")
y
```

Various extractors exist for `PJRTBuffer` objects, for example:

```{r}
shape(x)
device(x)
```

We can multiply the two buffers with each other by passing them to the `pjrt_execute()` function along with the executable.

```{r}
out <- pjrt_execute(executable, x, y)
out
```

Conversion back to R is possible via the `as_array()` function.

```{r}
as_array(out)
```

A noteworthy aspect of PJRT is that its execution model is fundamentally asynchronous: `pjrt_execute()` returns immediately with `PJRTBuffer` objects whose contents may not yet be available, and dependencies between successive calls are tracked internally so that subsequent operations can be enqueued without waiting for previous ones to finish.
The \CRANpkg{pjrt} package surfaces this via `is_ready()` for non-blocking status checks, `await()` to block until a buffer is ready, and `as_array_async()` together with `value()` to overlap device-to-host transfers with other host-side work.
This makes it possible to pipeline computation, communication, and host-side R code, which is particularly valuable on accelerators where launch latency would otherwise dominate.
Conversion back to R via `as_array()` implicitly synchronizes, since the result has to be materialized.
This async execution model carries through to \CRANpkg{anvl}: because `AnvlArray`s on the XLA backend wrap `PJRTBuffer`s, calls to `jit()`-compiled \CRANpkg{anvl} functions return immediately with `AnvlArray`s whose contents may not yet be ready, and the same `is_ready()`, `await()`, and `as_array_async()` interface applies.

Because `PJRTBuffer` objects are external pointers, custom (de)serialization methods are needed.
For this, we implement the safetensors format [@ref-safetensors].

```{r}
library(safetensors)
tmp <- tempfile(fileext = ".safetensors")
safetensors::safe_save_file(list(x = out), tmp, framework = "pjrt")
reloaded <- safetensors::safe_load_file(tmp, framework = "pjrt")
reloaded$x
```

### \CRANpkg{quickr}

The \CRANpkg{quickr} package serves as a second, experimental, backend for \CRANpkg{anvl} and was independently developed.
It transpiles R code to Fortran code and compiles it.
Below, we can recreate the StableHLO program from above.
One semantic difference is that it uses data type `f64` instead of `f32` because \CRANpkg{quickr} operates directly on R as opposed to \CRANpkg{pjrt} which defines its own array type.

```{r}
library(quickr)

multiply_r <- function(x, y) {
  declare(
    type(x = double(NA, NA)),
    type(y = double(NA, NA))
  )
  x * y
}
f <- quick(multiply_r)
f(
  matrix(c(1, 2, 3, 4), nrow = 2),
  matrix(c(5, 6, 7, 8), nrow = 2)
)
```

There are also various other differences with respect to the underlying implementation.
Instead of offering a builder API like in \CRANpkg{stablehlo}, \CRANpkg{quickr} works via static analysis of the R code.
Furthermore, the argument types of the inputs need to be specified via the `declare()` function.
Here, `double(NA, NA)` means a 2-dimensional array with unknown dimensions, but concrete dimensions can also be specified.
One advantage of \CRANpkg{quickr} over \CRANpkg{pjrt} and \CRANpkg{stablehlo} is that the former does not necessitate recompilation for every new input type combination (which includes the shapes).
A disadvantage is that \CRANpkg{quickr} currently does not support GPU acceleration and also less data types.

We stress that the \CRANpkg{quickr} backend is currently experimental and not yet usable in practice: only a small subset of \CRANpkg{anvl}'s primitives have a `quickr` lowering rule, and the backend does not yet exploit \CRANpkg{quickr}'s native support for shape dynamism, so today it offers no real advantage over the XLA backend.
The long-term goal of the \CRANpkg{quickr} backend is to enable ahead-of-time (AOT) compilation with proper shape dynamism, so that a compiled program can be reused across a wide range of input shapes without re-tracing or recompilation.
Realizing this requires reworking the `AnvlGraph` intermediate representation so that its shape datatypes can carry *symbolic* shapes rather than only fully static ones, which is a substantial change to the IR and is still work in progress.

### Automatic Differentiation

Besides JIT compilation, another core feature of \CRANpkg{anvl} is its ability to automatically compute derivatives of R functions.
Here, we briefly cover some central ideas behind AD to provide the background for understanding \CRANpkg{anvl}'s design.
For a comprehensive treatment, we refer the reader to @ref-griewank2008evaluating.

#### Evaluation Procedures and the Chain Rule

AD exploits the fact that every computer program, no matter how complex, executes a sequence of elementary operations whose derivatives are known.
Following @ref-griewank2008evaluating, consider a function $F : \R^n \to \R^m$ that is evaluated by a computer program.
Any such evaluation can be decomposed into a sequence of elementary operations (called *elemental functions*) $\varphi_i$ such as addition, multiplication, $\sin$, $\exp$, etc.
The evaluation produces a sequence of intermediate variables

$$v_i = \varphi_i(v_j)_{j \prec i}, \qquad i = 1, \ldots, l$$

where $v_{1-n}, \ldots, v_0$ are the input variables (i.e., $x$), $v_1, \ldots, v_l$ are the intermediate variables, and $v_{l-m+1}, \ldots, v_l$ are the output variables (i.e., $y$).
Here, the notation $j \prec i$ indicates that $v_j$ is a direct argument of $\varphi_i$.
This sequence of assignments is called an *evaluation procedure* and can be visualized as a directed acyclic graph (the *computational graph*), where each node corresponds to an intermediate variable and edges represent data dependencies.

Since each elemental function $\varphi_i$ is differentiable, the chain rule can be applied systematically to compute derivatives of $F$.
The two main modes of AD differ in the *order* in which the chain rule is evaluated, leading to different computational trade-offs.

#### Forward Mode

Forward-mode AD propagates derivatives alongside the function evaluation, computing a Jacobian-vector product $\dot{y} = F'(x) \dot{x}$ in a single forward sweep.
Obtaining the full Jacobian $F'(x) \in \R^{m \times n}$ requires $n$ such sweeps, which makes forward mode efficient when $n$ is small relative to $m$.
\CRANpkg{anvl} currently only implements reverse mode, which we describe next.

#### Reverse Mode

Reverse-mode AD (also called the *adjoint* mode) propagates sensitivities backward through the evaluation procedure.
Given an adjoint vector $\bar{y} \in \R^m$, it computes for each intermediate variable $v_i$ the corresponding adjoint

$$\bar{v}_i = \sum_{k \succ i} \bar{v}_k \fp{v_k}{v_i}$$

proceeding in reverse order, from outputs to inputs, where $k \succ i$ denotes that $v_i$ is a direct argument of $\varphi_k$.
The final result $\bar{x} = \bar{y}^\top F'(x)$ is a *vector-Jacobian product* (VJP).

The key advantage of reverse mode is that for a scalar-valued function ($m = 1$), setting $\bar{y} = 1$ yields the full gradient $\nabla F(x) \in \R^n$ in a *single* reverse sweep, regardless of $n$.
This result is sometimes referred to as the "cheap gradient principle": the computational cost of evaluating the gradient is bounded by a small constant multiple (typically 2--5) of the cost of evaluating $F$ itself (@ref-griewank2008evaluating, Chapter 3).
This makes reverse mode dramatically more efficient than forward mode for the common case of gradient-based optimization, where one differentiates a scalar loss with respect to potentially many parameters.

The trade-off is that reverse mode requires access to the intermediate values $v_i$ from the forward evaluation during the backward sweep, since they are needed to evaluate the local partial derivatives $\fp{v_k}{v_i}$.
This introduces additional memory overhead that scales with the depth of the computation.

In \CRANpkg{anvl}, we implement backward-mode AD as a graph-to-graph transformation, as described in the next section.
Each primitive operation stores its own local VJP rule, and the transformation walks the computational graph in reverse topological order to accumulate gradients.

## Package Design and Features

Having coverved the underlying basics, we will now cover the main design of the \CRANpkg{anvl} frontend.
The core idea of the package is centered around code transformations, an approach adopted from JAX [@ref-jax].
In general, there are three types of transformations:

1. Transforming R code into a computational graph via tracing.
2. Transforming a computational graph into another computational graph (e.g. for automatic differentiation).
3. Transforming a computational graph into an executable (either via \CRANpkg{stablehlo} and \CRANpkg{pjrt} or \CRANpkg{quickr}).

We will now cover these three types of transformations in more detail.
Then, we explain how these transformations are wrapped in the main user interface.

### Code transformations

#### Tracing R code into a computational graph

When it comes to translating code into other code, there are in principle two approaches:

1. Static analysis of the code.
2. Dynamically executing the code with "special values" that record certain operations and build up a (partial) representation of the program.

In \CRANpkg{anvl}, we are following the latter approach, because it allows to seemlessly combine standard R code with jit-compilable operations.
The function that takes in an R function and returns a computational graph is called `trace_fn()`.
In order for this to be useful, the function needs to call into anvl primitive functions.
These primitive functions are similar to the stableHLO operations like `hlo_add` from earlier.

In the \CRANpkg{anvl} package these are called `nvl_<op>` functions.
Howveer, they are not exported directly, but in thin wrappers like `nv_add`, which perform some convenience transformations (like type casting) that make them more ergonomic to use.
These thin wrappers call into the primitive functions, however.
The primitive functions are essentially graph building functions, also similar to the `hlo_add` function from earlier.
Below, we create an R function that takes in two values and multiplies them.
Note that we could have also used the `*` infix operator, but we want to illustrate the use of the `nv_multiply` function.

```{r}
library(anvl)
multiply_r <- function(x, y) {
  nv_mul(x, y)
}
```

We can now trace this function into a computational graph.
At the beginning, this will create a `anvl::GraphDescriptor` that primitive functions can attach to to build up the graph.
The types for the inputs are inferred from the example arguments.
The `trace_fn` transformation uses these example inputs to obtain abstract `GraphBox` objects, which are passed around during the evaluation of the function.
They essentially define the type (shape and data type), and uniquely identify a specific `GraphValue` within a specific `GraphDescriptor`.
Note that `trace_fn` also takes in non-array values, which are evaluated like normal R code.

```{r}
graph <- trace_fn(multiply_r, list(x = nv_scalar(1, "f32"), y = nv_scalar(5, "f32")))
graph
```

Once the evaluation of `multiply_r` with the special `GraphBox` objects is complete, the `GraphDescriptor` is converted into a `Graph` object.
The only difference between them is that `GraphDescriptor` does some book-keeping during program creation which is not needed anymore afterwards.

From the printed output of the graph we can see that a graph has:

1. Inputs: `GraphValue`s that are the inputs to the graph.
2. Constants: `GraphValue`s that are constants in the graph.
3. Body: A list of `PrimitiveCall` objects, which represent the primitive operations that make up the graph.
4. Outputs: `GraphValue`s that are the outputs of the graph.
5. In-tree and Out-tree: Because `Graph`s only take in array values, but we want to support nested data strctures, the in-tree and out-tree define how to (un-)flatten the inputs and outputs of the graph.

STATIC SHAPES!!!

#### Transforming a Graph into another Graph

Once our program no longer consists arbitrary R code, but is in a standardized form, we can transform the program by creating another `Graph` object.
Currently, the only example for such a type of transformation is the gradient transformation, which is implemented in the `transform_gradient()` function.
It takes in a `Graph` and the names of the arguments to compute the gradient with respect to.

```{r}
gradient_graph <- transform_gradient(graph, wrt = c("x", "y"))
```

The \CRANpkg{anvl} package does not put any restrictions on *how* such a transformation is implemented.
Usually, however, such a transformation needs to store information for how to transform each primitive.
For this reason, the `AnvlPrimitive` objects have transformation rules attached to them, accessed via `[[<rule>]]`.
For example, the reverse rule for the addition primitive is implemented as follows:

```{r}
prim_add[["reverse"]]
```

The transformation function `transform_gradient` basically walks the graph backwards and applies the reverse rules to each primitive, producing a `Graph` that contains the forward and backward pass.

#### The Gradient Transformation

The `transform_gradient()` function implements backward-mode automatic differentiation (also known as reverse-mode AD).
Given a `Graph` representing a function $f : \mathbb{R}^n \to \mathbb{R}$, it produces a new `Graph` that computes both the original output and the gradients of the output with respect to specified inputs.

More concretely, consider a computational graph as a sequence of primitive operations $v_k = g_k(v_{\text{pa}(k)})$, where $v_{\text{pa}(k)}$ denotes the parent values (inputs) of operation $k$.
The goal is to compute $\bar{v}_i = \frac{\partial f}{\partial v_i}$ for each intermediate value $v_i$.
Starting from $\bar{f} = 1$, the backward pass walks the graph in reverse topological order and accumulates gradients via the chain rule:
$$\bar{v}_i = \sum_{k : i \in \text{pa}(k)} \bar{v}_k \cdot \frac{\partial g_k}{\partial v_i}$$
Each primitive's reverse rule computes the local partial derivatives $\frac{\partial g_k}{\partial v_i}$ and multiplies them with the incoming gradient $\bar{v}_k$.
This is the standard vector-Jacobian product (VJP) formulation of reverse-mode AD, which avoids materializing the full Jacobian matrices of intermediate operations.
Instead, each reverse rule directly computes the product of the incoming gradient vector with the local Jacobian, which is typically much cheaper and can often be expressed in closed form.

In the implementation, the transformation walks the graph in reverse order.
For each `PrimitiveCall`, it looks up the corresponding reverse rule stored in the primitive's `@rules[["reverse"]]` field.
This reverse rule receives the original inputs and outputs of the primitive call (from the forward pass) as well as the incoming gradient $\bar{v}_k$, and returns the gradient contributions for each input.
These contributions are accumulated across all uses of a value to obtain the total gradient $\bar{v}_i$.

Because the reverse rules are themselves expressed in terms of primitive operations (i.e., they call `nvl_*` functions), the gradient graph is built using the same tracing mechanism as the forward pass.
This means that the gradient computation is itself a `Graph` that can be further transformed or compiled --- for example, one could in principle compute higher-order derivatives by applying `transform_gradient()` again to the gradient graph.

An important design choice is that the reverse rules are stored per-primitive in the `@rules` field, rather than being hard-coded in the transformation.
This makes it straightforward to add differentiation support for new primitives (see Section \ref{extending}) and keeps the transformation logic generic.

#### Compiling a Graph

Whenever we are done transforming out `Graph` and are ready to execute it, we need to lower it.
We can convert a `Graph` into a stablehlo program via the `stablehlo()` function.
The output of this is a list containing:

1. The StableHLO program.
2. The constant buffers in the `Graph`.

```{r}
out <- stablehlo(graph)
repr(out[[1]])
out[[2]]
```

While stablehlo allows to embed constants in the program, this can considerably increase the program size (think of big arrays) and slow down compile times.
For this reason, non-scalar (TODO) constants are not inlined into the stableHLO executable, but added as inputs to the program and need to be provided when executing the program.

Next, we will describe how these transformations are wrapped in the main user interface.

### Main User Interface

The main user interface is centered around `jit()` and transformations functions like `gradient()` or `value_and_gradient()`.
Furthermore, we can create arrays, which we will cover first.

#### The `AnvlArray`

In order to create data for our programs, we can use the `nv_array()`, `nv_scalar()` and `nv_empty()` functions.
These essentially wrap the `pjrt_buffer()` and `pjrt_scalar()` functions from the \CRANpkg{pjrt} package.

```{r}
x <- nv_scalar(1, "f32")
x
y <- nv_array(1:4, "f32")
y
```

Just like the `PJRTBuffer` objects, `AnvlArray` objects can be serialized to a string and deserialized using the \CRANpkg{safetensors} R package.

#### `jit()`

`jit()` is the main function that takes in an R function and returns a JIT-compiled function
Below, we apply it to a simple R function that prints a message and then multiplies two values.
The print statement is included for didactive purposes and will become clear later.

```{r}
multiply_print_r <- function(x, y) {
  print("Hello useR!")
  nv_mul(x, x)
}
multiply_jit <- jit(multiply_print_r)
```

The output function is a regular R function, which contains the "recipe" for transforming the R function into a `Graph`, then lowering into a StableHLO program and finally executing it.

```{r}
multiply_jit(x, y)
```

Of course, compiling the function every time the function is called would be very inefficient.
Therefore, a cache is maintained that stores the executable (and constant values) for each unique combination of input types.

If we execute it again, we get the same output, but the print statement is not shown.

```{r}
multiply_jit(x, y)
```

If we apply it again to different input values (of the same types), we get a cache hit and directly execute the cached executable.

```{r}
multiply_jit(nv_scalar(2, "f32"), nv_scalar(3, "f32"))
```

Because the compiled executable contains only the array operation, the print statement is not shown.
If we apply the function to values of different types, we get a cache miss and the function is recompiled.
This will show the print message again.

```{r}
multiply_jit(nv_scalar(1, "i32"), nv_scalar(5, "i32"))
```

We can also JIT-compile functions that take in lists of arrays and output lists of arrays.

```{r}
multiply_jit_nested <- jit(function(lst) {
  list(nv_mul(lst[[1]], lst[[2]]))
})
multiply_jit_nested(list(x, y))
```

This will use the `in_tree` and `out_tree` fields of the `Graph` to flatten and unflatten the list of arrays.

The functions we JIT-compile can also take in so called "static" arguemnts, that can be standard R values.
Below, we define a function that either multiplies or adds two values, depending on the value of the `op` argument.
We need to mark the `op` argument as static.

```{r}
add_or_multiply <- jit(function(x, y, op) {
  if (op == "add") {
    nv_add(x, y)
  } else {
    nv_mul(x, y)
  }
}, static = "op")
```

The difference between static and non-static arguments is that for the former the function is recompiled for ever unique input type combination, while for the latter it needs to be recompiled for every unique value of the static argument.

```{r}
add_or_multiply(x, y, "add")
add_or_multiply(x, y, "multiply")
```

#### Transformations like `gradient()`

Just like `jit()`, the `gradient()` function takes in a function and outputs a function that contains the recipe for transforming the input function into a `Graph` representing the gradient of the input function.

```{r}
add_fn <- function(x, y) {
  nv_add(x, y)
}
grad <- gradient(add_fn, wrt = c("x", "y"))
```

In order to be able to execute the gradient function, we need to wrap it in a `jit()` call.
We can also jit a function that contains "standard" anvl operations and calls into the gradient function.

```{r}
x <- nv_scalar(1, "f32")
y <- nv_scalar(5, "f32")
f <- function(x, y) {
  z <- nv_add(x, y)
  grads <- grad(x, z)
  grads
}
```

The way this works is the following:
When tracing the `val_and_grad` function, a `GraphDescriptor` is first initialized.
When the `add_or_multiply` function is called, it populates the `GraphDescriptor` with the inputs `x` and `y` and adds the `PrimitiveCall` to the body representing the multiplication of `x` and `y`.
When we then call into the `grad` function, we create another `GraphDescriptor`, because we need a graph-representation of the of the `add_or_multiply` function so we can transform it into its gradient.
After this is done, the resulting gradient graph is inlined into the parent `GraphDescriptor`.

We can see the output of this below:

```{r}
graph <- trace_fn(f, list(x = x, y = y))
graph
```

Because this makes `f` just another graph-building function, we can wrap it in a `jit()` call and execute it.

```{r}
f_jit <- jit(f)
f_jit(x, y)
```

#### JIT vs. Eager Mode

In addition to the JIT-compiled execution path described above, \CRANpkg{anvl} also supports an *eager* execution mode in which primitive operations are dispatched and executed immediately rather than being recorded into a `Graph`.
Concretely, calling `nv_add(1, nv_array(1:2))` outside of any tracing context performs the addition right away and returns an `AnvlArray`, instead of merely registering a `PrimitiveCall` on a `GraphDescriptor`.
Eager mode was introduced for two reasons.
First, it makes programs significantly easier to debug: intermediate values can be inspected directly with R tooling (e.g., `print()`, `browser()`), without having to disable JIT compilation or insert auxiliary `nv_print()` calls into a traced program.
Second, it removes the need to maintain two versions of every user-defined function --- one written in plain R for interactive use or testing, and another wrapped in `jit()` for execution.
With eager mode, the same R function can be called directly during development and only wrapped in `jit()` when the user is ready to compile it, since both modes share the same semantics for the underlying `nv_*` operations.
Eager mode is implemented as a special case of the same machinery used for tracing: when a primitive is invoked outside of any active `GraphDescriptor`, it falls back to executing immediately on the currently selected backend.

### Limitations of XLA

While the JIT-compilation approach is very powerful, it inherits a few restrictions from the XLA compiler that shape how programs must be written.

The most prominent one is that all shapes must be known at compile time.
While the StableHLO language itself does support dynamic shapes, the XLA compiler requires them to be fully static, which also allows it to apply more aggressive optimizations.
As a consequence, operations whose output shape depends on the values of the inputs, such as `unique()` or boolean masking, cannot be expressed directly.
A common workaround is to pad inputs to a fixed maximum size and mask out the irrelevant entries with e.g. `nv_where()`.
This preserves static shapes at the cost of some wasted computation.
On GPUs this extra work is often acceptable because idle lanes are cheap, but on CPUs padding tends to hurt performance much more noticeably.

A related restriction is that StableHLO operations are designed so that every precondition can be verified at compile time, rather than producing runtime errors.
For example, subsetting-like operations such as `nvl_dynamic_slice()` silently clamp the slice indices to the valid range instead of raising an out-of-bounds error, and in `nvl_gather()` out-of-bounds indices yield an implementation-defined result.
This means that any argument whose value must be validated (e.g. an index vector that must lie within a given range) effectively has to be a static argument to the jitted function, so that \CRANpkg{anvl} can check it during tracing rather than relying on the compiled program to do so.
In practice, this is usually not a severe restriction, but it is something users need to be aware of when designing their programs.


## Examples

TODO: HMC example.

## Extending

One of the key design goals of \CRANpkg{anvl} is to be extendable.
It is possible to add new primitive operations, transformations, and even backends.

### New Primitive Operations

The simplest way to extend \CRANpkg{anvl} is to add new primitive operations via `new_primitive()`.
A primitive bundles an abstract-evaluation rule (mapping input shapes and dtypes to output shapes and dtypes) with a set of per-transformation rules — a StableHLO lowering rule for the XLA backend, a `quickr` lowering rule for the Fortran backend, and a reverse-mode rule for automatic differentiation.
Any operation that can be expressed in terms of StableHLO and \CRANpkg{quickr} primitives can be added this way, and the same mechanism is used for all primitives shipped with \CRANpkg{anvl} itself.
For operations that cannot be expressed in StableHLO --- for example when wrapping a hand-written CUDA kernel or an existing C/C++ routine --- the XLA backend additionally exposes PJRT's custom-call FFI mechanism via `pjrt::pjrt_register_custom_call()`, which lets users dispatch to a C/C++ handler from inside a compiled program (with separate CPU and CUDA implementations selected per PJRT platform).
A worked end-to-end example, including how to write the rules and register a custom call, is given in the [extending primitives vignette](https://r-xla.github.io/anvl/articles/extending_primitive.html).

### New Transformations

Because the `Graph` representation is a simple data structure consisting of inputs, outputs, and a sequence of `PrimitiveCall` objects, it is straightforward to implement new graph-to-graph transformations.
The `transform_gradient()` function, which implements backward-mode automatic differentiation, is the primary example of such a transformation.
It walks the graph in reverse order and applies the backward rule stored in each primitive's `@rules[["backward"]]` field to accumulate gradients.

Implementing a new transformation follows the same pattern: iterate over the `PrimitiveCall` objects in the graph, look up a corresponding rule for each primitive, and use these rules to construct a new graph.
For example, one could implement forward-mode automatic differentiation by walking the graph *forward* and propagating Jacobian-vector products, or implement a graph optimization pass that fuses or simplifies sequences of primitive calls.
The key design decision is that the \CRANpkg{anvl} package does not impose any constraints on how a transformation is implemented — it only provides the `Primitive` rule mechanism as a convenient way to store per-operation information.

### Custom Backends

Backends are represented by `AnvlBackend` objects that bundle the operations \CRANpkg{anvl} needs in order to treat a storage-and-execution mechanism as a target.
In essence, an `AnvlBackend` is a small collection of functions covering data construction (`new_data`), metadata accessors (`dtype`, `shape`, `ambiguous`), conversions (`as_array`, `as_raw`), device handling (`platform`, `device`, `new_device`), printing (`print_data`), and JIT compilation (`jit`).
Each backend is registered under a string name and can be selected via `local_backend()` or `with_backend()`.
Two backends are currently shipped with \CRANpkg{anvl}: the default `"xla"` backend, which wraps `PJRTBuffer` objects and lowers `Graph`s to StableHLO, and the `"quickr"` backend described earlier, which stores plain R arrays and compiles `Graph`s to R code via \CRANpkg{quickr}.
Adding a new backend consists of implementing the `AnvlBackend` interface and, for each primitive that the backend supports, providing a lowering rule under a new name in the primitive's `@rules` field.

Conceptually, the `AnvlBackend` interface plays the role of a "mini PJRT": it captures the essential notions that any backend must provide --- how to hold array data, how to describe it, how to move it around, and how to compile and execute a graph --- but avoids the complexity of the full PJRT C API.
This makes it considerably easier to plug in new execution strategies that do not fit the PJRT model.
For example, a pure-R backend that simply interprets `Graph`s using base-R array operations can be implemented as an `AnvlBackend` without touching a compiler toolchain or a C++ plugin.

That said, the `AnvlBackend` interface is still work in progress and not yet fully specified.
Its precise shape will continue to evolve as we gain experience with additional backends, so authors of custom backends should currently expect occasional breaking changes.
A more streamlined and well-documented interface for registering custom backends is an area we plan to improve.

## Testing

Ensuring correctness of both the forward evaluation and the gradient computation is critical for a code transformation framework.
In \CRANpkg{anvl}, we bootstrap correctness by comparing against the \CRANpkg{torch} package wherever possible.

For forward-pass testing, each primitive's JIT-compiled output is compared against the corresponding \CRANpkg{torch} function on randomly generated inputs.
For example, to test `nvl_add`, we generate random arrays, compute the result using both `jit(nvl_add)(x, y)` and `torch::torch_add(x, y)`, and assert that the outputs match up to numerical tolerance.
This is done using helper functions like `expect_jit_torch_binary()` and `expect_jit_torch_unary()`, which handle the data generation and comparison.

For backward-pass testing, the same strategy is applied to gradient computation.
The anvl gradient (obtained via `jit(gradient(f))(x)`) is compared against the gradient computed by \CRANpkg{torch}'s autograd engine.
For non-scalar functions, both sides are wrapped with a summation to produce a scalar loss before differentiating, ensuring a consistent comparison.

Because \CRANpkg{torch} is not a hard dependency of the package, the torch-based tests are stored in `inst/extra-tests/` and are conditionally sourced only when \CRANpkg{torch} is installed.
This keeps the test suite lightweight for users who do not have \CRANpkg{torch}, while still running the full comparison suite in CI.
For primitives that have no direct \CRANpkg{torch} equivalent (e.g., random number generation or bitwise operations), manual tests with hardcoded expected values are used instead.

## Benchmarks

TODO: HMC benchmark.

Runtime benchmarks comparing \CRANpkg{anvl} on CPU and GPU are available at \url{https://r-xla.github.io/benchmarks/}.

## Discussion

<!-- Summary -->
In this paper, we introduced the \CRANpkg{anvl} package, which is a code transformation framework for the R language.
Execution is handled via JIT compilation through either the OpenXLA compiler framework or by transpiling to Fortran via \CRANpkg{quickr}.
When using the OpenXLA backend, it is possible to make use of accelerators such as GPUs.
Furthermore, the package supports first-order automatic differentiation.

<!-- Limitations -->
The biggest limitations of the current approach are twofold:

a) Programs need to have static shapes, which means that operations such as `unique()` are currently not supported.
b) For the OpenXLA backend, programs need to be re-compiled for each new unique input type combination.

TODO: Tracing vs. Static analysis.

<!-- Future work -->
### Future Work

We group the directions for future work into five categories: lifting current limitations and improving the IR, filling gaps in the API, adding more code transformations, supporting more compute platforms, and building downstream applications on top of \CRANpkg{anvl}.

#### Lifting Current Limitations and IR Improvements

Most of the limitations discussed above stem from the design of the intermediate `AnvlGraph` and from XLA-imposed restrictions on the StableHLO programs we emit.
The most pressing direction is to extend the IR datatypes underlying `AnvlGraph` (in particular `AbstractArray`) to carry symbolic shapes rather than only fully static ones, so that the same graph can describe a family of input shapes.
This is a prerequisite for either of the two routes that would lift the static-shape and recompilation restrictions.
The first is to swap out the XLA compiler for the IREE compiler [@ref-iree], which also consumes StableHLO programs but natively supports dynamic shapes; the downside is that IREE is considerably less mature than XLA and not yet a practical drop-in.
The second is to extend the \CRANpkg{quickr} backend, which targets Fortran and does support shape dynamism, so that it covers a larger fraction of \CRANpkg{anvl}'s primitives and would lift the second restriction as well (at least on CPU); today this backend is not yet far enough along to serve as such a replacement.
A related improvement is to decouple `AnvlGraph` from StableHLO more thoroughly: currently the IR is tied too closely to StableHLO, which limits the ability to target alternative backends and to represent higher-level optimizations that do not map directly to StableHLO concepts.

#### Filling Gaps in the API

The set of primitive operations currently supported by \CRANpkg{anvl} and \CRANpkg{stablehlo} is rather narrow and is the simplest aspect of the package to extend.
On top of new primitives, additional API functions can be added, in particular focusing on linear algebra operations that are common in statistical and machine learning workloads.

#### Adding More Code Transformations

So far, the only graph-to-graph transformation \CRANpkg{anvl} ships is reverse-mode automatic differentiation.
Several other transformations would substantially extend the package's capabilities: forward-mode automatic differentiation, which together with reverse mode enables efficient computation of higher-order derivatives such as Jacobians and Hessians; improved vectorization transformations in the spirit of JAX's `vmap`; and JET-style static analyzers that walk an `AnvlGraph` to catch shape, dtype, or other correctness issues before compilation.

#### Supporting More Compute Platforms

On the hardware side, the set of supported PJRT plugins could be expanded.
The ZML project [@ref-zml] ships PJRT plugins for ROCm (AMD GPUs) and CUDA on Arm, and adopting these would extend \CRANpkg{anvl}'s reach to these platforms with comparatively little effort.
Proper support for Apple Silicon GPUs is more uncertain: the official `jax-metal` plugin appears to no longer be actively developed, but the experimental [jax-mps](https://github.com/tillahoffmann/jax-mps) project explores routing JAX through Metal Performance Shaders, and if it matures it could provide a path to first-class Mac GPU execution for \CRANpkg{anvl} as well.
The main blocker for supporting additional platforms is not integration effort but testing infrastructure: each backend needs continuous-integration runners with the corresponding hardware in order to keep the test suite green, and access to such runners (especially for AMD GPUs, Arm, and Apple Silicon) is currently the limiting factor.

#### Downstream Applications

\CRANpkg{anvl} is designed as low-level infrastructure on which more specialized libraries can be built.
We see several high-impact directions:

- A Bayesian inference library along the lines of NumPyro [@ref-numpyro] or BlackJAX [@ref-blackjax], including HMC and variational inference algorithms that benefit directly from JIT compilation, GPU acceleration, and AD.
- A deep learning library exposing high-level neural network modules and gradient-based optimizers, analogous to the Flax/Optax stack in the JAX ecosystem.
- Differential equation solvers (analogous to Diffrax), which combine naturally with AD for gradient-based fitting of dynamical systems.
- Reimplementing statistical methods that already exist in R but do not scale to large datasets because their current implementations are too slow, where JIT compilation, GPU acceleration, and AD could yield substantial speedups while keeping the implementation in pure R.

## Acknowledgments

Sebastian Fischer is supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – 460135501 (NFDI project MaRDI).


## Appendix {.appendix}

### AI Usage

AI coding agents were used for the development of the software.
However, all AI-generated code was reviewed by the authors.
